---
title: "Untitled"
output: html_document
date: '2025-04-28'
---

This is a script written by Jens Bager Christensen (jbc53@cam.ac.uk) to evaluate our primary human NEP cultures using scRNA-seq. 

Packages are loaded
```{r}
library(Seurat)
library(ggplot2)
library(tidyverse)
#library(reticulate)
#library(leiden)
```

Functions are defined
```{r}
#set seed for consistency
set.seed(1234)

#Defining a function, which clusters a RDS in a range of resolutions, and calculates both the silhouette score and the negative silhouette proportion to determine which resolution is optimal
clusteringKit <- function(name, assay, dim, res.range, reduction){
  DefaultAssay(name) <- assay
  #using FindNeighbors (Seurat), a KNN based clustering algorithm, to compute a SNN graph, which then can be used to cluster the cells w. their neighbors
  name <- FindNeighbors(name, dims=1:dim, reduction=reduction)
  for (i in res.range){
    #FindClusters is used to cluster the aforementioned SNN graph by optimising the modularity function. This is carried out for all resolutions in the range of interest.
    name <- FindClusters(name, resolution=i)
  }
DefaultAssay(name) <- assay
  dist.matrix <- dist(x = Embeddings(object = name[[reduction]])[, 1:dim])
 clusters <- paste0(assay,"_snn_res.", res.range)
  getSil <- function(clr) {
    clr <- name@meta.data[[clr]]
    sil <- cluster::silhouette(x = as.numeric(x = as.factor(x = clr)), dist = dist.matrix)
    sil_value <- sil
    return(sil_value)
  }
  
  sls <- lapply(clusters, getSil) #mc.cores = 1)
  sls_median <- sapply(sls, function(x) median(x[,3])) %>% setNames(., res.range)
  sls_neg_prop <- sapply(sls, function(x) sum(x[,3]<0)/length(x[,3])) %>% setNames(., res.range)
  p_list <- lapply(res.range, function(res){
    #Idents(name) = name@meta.data[paste0(assay, "_snn_res.",res)]
    DimPlot(name, reduction = "umap", label = TRUE, group.by = paste0(assay, "_snn_res.",res), pt.size =0.3, raster=F)
  })
  return(list(name=name, sls_median=sls_median, sls_neg_prop=sls_neg_prop, dimplots=p_list))
}
```

Read in the output of CellRanger aligned to GRCh38 and create a RDS object out of this. 
Classify cells based on the multiplex assay, which contains expression of the 12 used CMOs.
Thereafter, remove CMO311 and CMO312 because these samples were mouse NEPs.
Following this vignette : https://www.10xgenomics.com/analysis-guides/tag-assignment-of-10x-Genomics-cellplex-data-using-seurats-htodemux-function
```{r}
#read in the output from 'cellranger multi' performed using CellRanger 8.0.1. Aligning only to GRCh38
data <- Seurat::Read10X("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/primary_human_NEPs_GRCh38/outs/multi/count/raw_feature_bc_matrix/")

#Add the CellPlex as a CMO assay
adata = CreateSeuratObject(counts = data$`Gene Expression`)
adata[['CMO']] = CreateAssayObject(counts = data$`Multiplexing Capture`)

#identify cells with confident sample called based on cellranger. 
library(data.table)
cells <- fread("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/primary_human_NEPs_GRCh38/outs/multi/multiplexing_analysis/assignment_confidence_table.csv")

#remove cells assigned either Multiplet or Unassigned
cells <- cells %>% filter(Assignment != "Multiplet" & Assignment != "Unassigned")

#extract only barcodes called with confident sample call
adata <- subset(adata, cells = cells$Barcode)

#remove cell without CMO counts
DefaultAssay(adata)<-"CMO"
adata <-  subset(x = adata, subset = nCount_CMO > 0)

#Normalize the CMO data and run the HTODemux function. Read more https://satijalab.org/seurat/archive/v3.1/hashing_vignette.html
adata <- NormalizeData(adata, assay = "CMO", normalization.method = "CLR")
adata <- HTODemux(adata, assay = "CMO", positive.quantile = 0.99)

#print out the amount of negative, doublets and singlets
print(table(adata$CMO_classification.global))

#Check the expression levels of CMO ID
#The true positive cells should be well distanced from the other CMO IDs 
RidgePlot(adata, assay = "CMO", features = rownames(adata[["CMO"]])[1:12], ncol = 6)

#include only CMO301-CMO310 because these are human samples. Thus, doublets are also discarded here
Idents(adata) <- "hash.ID"
adata_hu <- subset(adata, idents = c("CMO301", "CMO302", "CMO303", "CMO304", "CMO305",
                                  "CMO306", "CMO307", "CMO308", "CMO309", "CMO310"))

#add the BRC and HDBR number and GSW to the meta data
adata_hu@meta.data <- adata_hu@meta.data %>%
  transform(sample = case_when(
    hash.ID == "CMO301" ~ "BRC_2584 (9w)",
    hash.ID == "CMO302" ~ "BRC_2583 (9w_4d)",
    hash.ID == "CMO303" ~ "BRC_2591 (10w_5d)",
    hash.ID == "CMO304" ~ "BRC_2593 (12w)",
    hash.ID == "CMO305" ~ "HDBR_18265 (12w)",
    hash.ID == "CMO306" ~ "HDBR_18294 (12w)",
    hash.ID == "CMO307" ~ "HDBR_18267 (13w)",
    hash.ID == "CMO308" ~ "HDBR_18285 (13w)",
    hash.ID == "CMO309" ~ "HDBR_18290 (15w)",
    hash.ID == "CMO310" ~ "HDBR_18275 (17w)",
    ))

#add the a joint time meta data variable
adata_hu@meta.data <- adata_hu@meta.data %>%
  transform(GSW = case_when(
    hash.ID == "CMO301" ~ "9w",
    hash.ID == "CMO302" ~ "9w",
    hash.ID == "CMO303" ~ "10w",
    hash.ID == "CMO304" ~ "12w",
    hash.ID == "CMO305" ~ "12w",
    hash.ID == "CMO306" ~ "12w",
    hash.ID == "CMO307" ~ "13w",
    hash.ID == "CMO308" ~ "13w",
    hash.ID == "CMO309" ~ "15w",
    hash.ID == "CMO310" ~ "17w",
    ))

#add the a sex meta data variable based on c("DDX3Y", "KDM5D", "RPS4Y1") expression
adata_hu@meta.data <- adata_hu@meta.data %>%
  transform(sex = case_when(
    sample == "BRC_2584 (9w)" ~ "XX",
    sample == "BRC_2583 (9w_4d)" ~ "XX",
    sample == "BRC_2591 (10w_5d)" ~ "XY",
    sample == "BRC_2593 (12w)" ~ "XY",
    sample == "HDBR_18265 (12w)" ~ "XX",
    sample == "HDBR_18294 (12w)" ~ "XX",
    sample == "HDBR_18267 (13w)" ~ "XY",
    sample == "HDBR_18285 (13w)" ~ "XX",
    sample == "HDBR_18290 (15w)" ~ "XY",
    sample == "HDBR_18275 (17w)" ~ "XY",
    ))
```

_loading velocyte outputs and modifying cell ID to match the cell ranger output_
```{r}
library(SeuratWrappers)
BRC_2583 <- ReadVelocity("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/Velocyto_to_GRCh38/BRC_2583/velocyte_BRC_2583_GRCh38.loom")
BRC_2584 <- ReadVelocity("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/Velocyto_to_GRCh38/BRC_2584/velocyte_BRC_2584_GRCh38.loom")
BRC_2591 <- ReadVelocity("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/Velocyto_to_GRCh38/BRC_2591/velocyte_BRC_2591_GRCh38.loom")
BRC_2593 <- ReadVelocity("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/Velocyto_to_GRCh38/BRC_2593/velocyte_BRC_2593_GRCh38.loom")
HDBR_18265 <- ReadVelocity("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/Velocyto_to_GRCh38/HDBR_18265/velocyte_HDBR_18265_GRCh38.loom")
HDBR_18267 <- ReadVelocity("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/Velocyto_to_GRCh38/HDBR_18267/velocyte_HDBR_18267_GRCh38.loom")
HDBR_18275 <- ReadVelocity("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/Velocyto_to_GRCh38/HDBR_18275/velocyte_HDBR_18275_GRCh38.loom")
HDBR_18285 <- ReadVelocity("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/Velocyto_to_GRCh38/HDBR_18285/velocyte_HDBR_18285_GRCh38.loom")
HDBR_18290 <- ReadVelocity("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/Velocyto_to_GRCh38/HDBR_18290/velocyte_HDBR_18290_GRCh38.loom")
HDBR_18294 <- ReadVelocity("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/Velocyto_to_GRCh38/HDBR_18294/velocyte_HDBR_18294_GRCh38.loom")

#change cell ID for all samples
#BRC_2583
colnames(BRC_2583$spliced) <- gsub(".*:","",colnames(BRC_2583$spliced))
colnames(BRC_2583$spliced) <- gsub("x","-1",colnames(BRC_2583$spliced))
colnames(BRC_2583$unspliced) <- gsub(".*:","",colnames(BRC_2583$unspliced))
colnames(BRC_2583$unspliced) <- gsub("x","-1",colnames(BRC_2583$unspliced))
#BRC_2584
colnames(BRC_2584$spliced) <- gsub(".*:","",colnames(BRC_2584$spliced))
colnames(BRC_2584$spliced) <- gsub("x","-1",colnames(BRC_2584$spliced))
colnames(BRC_2584$unspliced) <- gsub(".*:","",colnames(BRC_2584$unspliced))
colnames(BRC_2584$unspliced) <- gsub("x","-1",colnames(BRC_2584$unspliced))
#BRC_2591
colnames(BRC_2591$spliced) <- gsub(".*:","",colnames(BRC_2591$spliced))
colnames(BRC_2591$spliced) <- gsub("x","-1",colnames(BRC_2591$spliced))
colnames(BRC_2591$unspliced) <- gsub(".*:","",colnames(BRC_2591$unspliced))
colnames(BRC_2591$unspliced) <- gsub("x","-1",colnames(BRC_2591$unspliced))
#BRC_2593
colnames(BRC_2593$spliced) <- gsub(".*:","",colnames(BRC_2593$spliced))
colnames(BRC_2593$spliced) <- gsub("x","-1",colnames(BRC_2593$spliced))
colnames(BRC_2593$unspliced) <- gsub(".*:","",colnames(BRC_2593$unspliced))
colnames(BRC_2593$unspliced) <- gsub("x","-1",colnames(BRC_2593$unspliced))
#HDBR_18265
colnames(HDBR_18265$spliced) <- gsub(".*:","",colnames(HDBR_18265$spliced))
colnames(HDBR_18265$spliced) <- gsub("x","-1",colnames(HDBR_18265$spliced))
colnames(HDBR_18265$unspliced) <- gsub(".*:","",colnames(HDBR_18265$unspliced))
colnames(HDBR_18265$unspliced) <- gsub("x","-1",colnames(HDBR_18265$unspliced))
#HDBR_18267
colnames(HDBR_18267$spliced) <- gsub(".*:","",colnames(HDBR_18267$spliced))
colnames(HDBR_18267$spliced) <- gsub("x","-1",colnames(HDBR_18267$spliced))
colnames(HDBR_18267$unspliced) <- gsub(".*:","",colnames(HDBR_18267$unspliced))
colnames(HDBR_18267$unspliced) <- gsub("x","-1",colnames(HDBR_18267$unspliced))
#HDBR_18275
colnames(HDBR_18275$spliced) <- gsub(".*:","",colnames(HDBR_18275$spliced))
colnames(HDBR_18275$spliced) <- gsub("x","-1",colnames(HDBR_18275$spliced))
colnames(HDBR_18275$unspliced) <- gsub(".*:","",colnames(HDBR_18275$unspliced))
colnames(HDBR_18275$unspliced) <- gsub("x","-1",colnames(HDBR_18275$unspliced))
#HDBR_18285
colnames(HDBR_18285$spliced) <- gsub(".*:","",colnames(HDBR_18285$spliced))
colnames(HDBR_18285$spliced) <- gsub("x","-1",colnames(HDBR_18285$spliced))
colnames(HDBR_18285$unspliced) <- gsub(".*:","",colnames(HDBR_18285$unspliced))
colnames(HDBR_18285$unspliced) <- gsub("x","-1",colnames(HDBR_18285$unspliced))
#HDBR_18290
colnames(HDBR_18290$spliced) <- gsub(".*:","",colnames(HDBR_18290$spliced))
colnames(HDBR_18290$spliced) <- gsub("x","-1",colnames(HDBR_18290$spliced))
colnames(HDBR_18290$unspliced) <- gsub(".*:","",colnames(HDBR_18290$unspliced))
colnames(HDBR_18290$unspliced) <- gsub("x","-1",colnames(HDBR_18290$unspliced))
#HDBR_18294
colnames(HDBR_18294$spliced) <- gsub(".*:","",colnames(HDBR_18294$spliced))
colnames(HDBR_18294$spliced) <- gsub("x","-1",colnames(HDBR_18294$spliced))
colnames(HDBR_18294$unspliced) <- gsub(".*:","",colnames(HDBR_18294$unspliced))
colnames(HDBR_18294$unspliced) <- gsub("x","-1",colnames(HDBR_18294$unspliced))

#remove duplicates
#BRC_2583
BRC_2583$spliced <- BRC_2583$spliced[!duplicated(rownames(BRC_2583$spliced)),]
BRC_2583$unspliced <- BRC_2583$unspliced[!duplicated(rownames(BRC_2583$unspliced)),]
#BRC_2584
BRC_2584$spliced <- BRC_2584$spliced[!duplicated(rownames(BRC_2584$spliced)),]
BRC_2584$unspliced <- BRC_2584$unspliced[!duplicated(rownames(BRC_2584$unspliced)),]
#BRC_2591
BRC_2591$spliced <- BRC_2591$spliced[!duplicated(rownames(BRC_2591$spliced)),]
BRC_2591$unspliced <- BRC_2591$unspliced[!duplicated(rownames(BRC_2591$unspliced)),]
#BRC_2593
BRC_2593$spliced <- BRC_2593$spliced[!duplicated(rownames(BRC_2593$spliced)),]
BRC_2593$unspliced <- BRC_2593$unspliced[!duplicated(rownames(BRC_2593$unspliced)),]
#HDBR_18265
HDBR_18265$spliced <- HDBR_18265$spliced[!duplicated(rownames(HDBR_18265$spliced)),]
HDBR_18265$unspliced <- HDBR_18265$unspliced[!duplicated(rownames(HDBR_18265$unspliced)),]
#HDBR_18267
HDBR_18267$spliced <- HDBR_18267$spliced[!duplicated(rownames(HDBR_18267$spliced)),]
HDBR_18267$unspliced <- HDBR_18267$unspliced[!duplicated(rownames(HDBR_18267$unspliced)),]
#HDBR_18275
HDBR_18275$spliced <- HDBR_18275$spliced[!duplicated(rownames(HDBR_18275$spliced)),]
HDBR_18275$unspliced <- HDBR_18275$unspliced[!duplicated(rownames(HDBR_18275$unspliced)),]
#HDBR_18285
HDBR_18285$spliced <- HDBR_18285$spliced[!duplicated(rownames(HDBR_18285$spliced)),]
HDBR_18285$unspliced <- HDBR_18285$unspliced[!duplicated(rownames(HDBR_18285$unspliced)),]
#HDBR_18290
HDBR_18290$spliced <- HDBR_18290$spliced[!duplicated(rownames(HDBR_18290$spliced)),]
HDBR_18290$unspliced <- HDBR_18290$unspliced[!duplicated(rownames(HDBR_18290$unspliced)),]
#HDBR_18294
HDBR_18294$spliced <- HDBR_18294$spliced[!duplicated(rownames(HDBR_18294$spliced)),]
HDBR_18294$unspliced <- HDBR_18294$unspliced[!duplicated(rownames(HDBR_18294$unspliced)),]
```

_for further analysis include only cells that have both reads from the cellranger and the velocyte output_
```{r}
#create a surrogate dgCMatrix object
loom_data <- BRC_2583

#change the surrogate dgCMatrix with the compined data from the 10 human samples
loom_data$spliced <- cbind(BRC_2583$spliced, BRC_2584$spliced, BRC_2591$spliced, BRC_2593$spliced, HDBR_18265$spliced, HDBR_18267$spliced, HDBR_18275$spliced, HDBR_18285$spliced, HDBR_18290$spliced, HDBR_18294$spliced)
loom_data$unspliced <- cbind(BRC_2583$unspliced, BRC_2584$unspliced, BRC_2591$unspliced, BRC_2593$unspliced, HDBR_18265$unspliced, HDBR_18267$unspliced, HDBR_18275$unspliced, HDBR_18285$unspliced, HDBR_18290$unspliced, HDBR_18294$unspliced)

#make it a v3 seurat object to avoid an error
options(Seurat.object.assay.version = 'v3')

#create a seurat object with the spliced count matrix as an assay
vel <- Seurat::CreateSeuratObject(counts = loom_data$spliced,
                          assay = "spliced",
                          min.cells = 0,
                          min.features = 0,
                          names.field = 2,
                          names.delim = "\\-")

#add the unspliced count matrix as an assay to the RDS object
vel[["unspliced"]] <- SeuratObject::CreateAssayObject(counts = loom_data$unspliced)

#called cells will not be exactly matching, because one RDS is made from the output of cellranger, the other from the output of velocyte
#evaluate the amount of cells lost
paste0("The velocyte processed RDS object contains ", length(colnames(vel)), " cells")
paste0("Of these ", length(which(is.element(colnames(vel), colnames(adata_hu)))), " are in the human data")

#keep only cells in the velocity RDS object, that is also present in the human data
vel <- vel[, colnames(adata_hu)]

paste0("The cellranger processed RDS object contains ", length(colnames(adata_hu)), " cells")
paste0("Of these ", length(which(is.element(colnames(adata_hu), colnames(vel)))), " are also in the subsetted velocyte processed RDS")

#similarly, keep only cells in the human data RDS that is also in the velocity RDS
adata_hu <- adata_hu[, colnames(vel)]

#extract the assay data from the velocyte processed object - both the spliced and unspliced assay
unspliced.data <- GetAssayData(object =  vel[['unspliced']], slot = 'counts')
spliced.data <- GetAssayData(object =  vel[['spliced']], slot = 'counts')

#add the spliced and unspliced assays to the human RDS
adata_hu[["unspliced"]] <- CreateAssayObject(data = unspliced.data )
adata_hu[["spliced"]] <- CreateAssayObject(data = spliced.data )
```

_perform initial QC of cells_

```{r}
#change assay to RNA
adata_hu@active.assay <- "RNA"

# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
adata_hu[["percent.mt"]] <- PercentageFeatureSet(adata_hu, pattern = "^MT-")

# Visualize QC metrics as a violin plot
VlnPlot(adata_hu, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), group.by = "sample", ncol = 3, alpha = 0.05)

#remove outliers beased on amount of features detected and too high mitochondrial reads
adata_hu <- subset(adata_hu, subset = nFeature_RNA > 200 & nFeature_RNA < 7500 & percent.mt < 5)

#run sctransform, normalising features, counts and percent mitochondrial reads
options(future.globals.maxSize = 8000 * 1024^2)
adata_hu <- SCTransform(adata_hu, vars.to.regress = c("nFeature_RNA", "nCount_RNA", "percent.mt"), verbose = FALSE)

#perform dimension reduction using PCA and UMAP to visualise the data 
adata_hu <- RunPCA(adata_hu, verbose = FALSE)
adata_hu <- RunUMAP(adata_hu, dims = 1:30, verbose = FALSE, return.model = T)

#visualise the data on a DimPlot
DimPlot(adata_hu, group.by = "sample") & NoAxes()

#perform cell cycle scoring and regress out the difference between S and G2M
adata_hu <- CellCycleScoring(adata_hu, s.features = cc.genes$s.genes, g2m.features = cc.genes$g2m.genes)

#visualise the phase on a DimPlot
DimPlot(adata_hu, group.by = "Phase")

#regress out the difference between G2M and S using sctransform
adata_hu$CC.Difference <- adata_hu$S.Score - adata_hu$G2M.Score
adata_hu <- SCTransform(adata_hu, vars.to.regress = c("CC.Difference"))

#redo the dimension reduction
adata_hu <- RunPCA(adata_hu, verbose = FALSE)
adata_hu <- RunUMAP(adata_hu, dims = 1:30, verbose = FALSE, return.model = T)

#visualise the Dimplot with metadata variables of interest
DimPlot(adata_hu, group.by = "sample") & NoAxes()
DimPlot(adata_hu, group.by = "Phase") & NoAxes()
DimPlot(adata_hu, group.by = "GSW") & NoAxes()
DimPlot(adata_hu, group.by = "sex") & NoAxes()
```

_Perform clustering and cluster analysis to annotate the clusters_
```{r}
#try to cluster using a range of resolutions to identifying a meaningful resolution parameter
adata_hu <- clusteringKit(adata_hu, assay="SCT", dim=30, res.range=seq(0.1, 3, 0.1), reduction="pca")


#plot to find the higest silhouette score w. low negative silhouette proportion
plot(adata_hu$sls_median~seq(0.1,3,0.1), type = "l", ylab = "Silhouette score") & plot(adata_hu$sls_neg_prop~seq(0.1,3,0.1), type = "l", ylab = "Negative silhouette proportion score")
#inspect the different tested resolutions to identify the optimal value
#a resolution of 0.3 seems to be fitting, not overclustering.

#redefine the seurat object
adata_hu <- adata_hu$name

#Plot with the identified resolution and compare to the previous annotations
DimPlot(adata_hu, label = T, label.box = T, repel = T, group.by = "SCT_snn_res.0.3") + ggtitle("Human cells clustered with a resolution of 0.3") & NoLegend() & NoAxes()

#look at expression of genes on Y chromosome ("DDX3Y", "RPS4Y1", "UTY") or on X chromosome, but only expressed in females (XIST)
VlnPlot(adata_hu, features = c("DDX3Y", "RPS4Y1", "UTY", "XIST"), group.by = "sample", ncol = 4)

#write out a RDS
saveRDS(adata_hu, "/mnt/beegfs7/home3/bayin/jbc53/scRNA-seq_2023/Christensen_et_al_2024/data/rds/primary_human_NEPs.rds")
```

_understanding initial clustering_
```{r}
#calculate marker genes for the clusters by setting the assay to RNA and the idents to the clustering computed with the optimal resolution.
DefaultAssay(adata_hu) <- 'SCT'
Idents(adata_hu) <- "SCT_snn_res.0.3"
adata_hu <- PrepSCTFindMarkers(adata_hu, assay = "SCT", verbose = TRUE)
adata_hu_markers <- FindAllMarkers(adata_hu, only.pos = TRUE, min.pct = 0.10, logfc.threshold = 0.25, assay = "SCT")

#write out the marker genes
write.csv(adata_hu_markers, "Hs_in_vitro/Hs_in_vitro_marker_genes.csv")

#cluster 6 & 8 have low nCount_RNA and nFeature_RNA. Possibly damaged cells.
VlnPlot(adata_hu, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"))

DimPlot(adata_hu, group.by = "SCT_snn_res.0.3", label = T) & NoAxes()
DimPlot(adata_hu, group.by = "sample") & NoAxes()
DotPlot(adata_hu, features = c("COL11A1", "FGF13" ,"PDGFRA", "COL1A1", "PDGFRB", "SOX2", "NES",
                               "GDF10", "GFAP", "HOPX", "ASCL1", "AQP4", "SLC1A2", "CENPF", "TOP2A"), group.by = "SCT_snn_res.0.3")

DotPlot(adata_hu, features = c("COL11A1" ,"DCN" ,"COL1A1", "PDGFRA", "SOX2", "NES", "GDF10", "GFAP", "HOPX", "ASCL1", "DLL1", "AQP4", "SLC1A2", "MKI67", "TOP2A"), group.by = "SCT_snn_res.0.3")

adata_hu_markers %>% filter(gene == "COL11A1")
adata_hu_markers %>% filter(cluster == 7)
FeaturePlot(adata_hu, "SOX2")
#0 COL11A1 / FGF13 / DCN (mesenchymal cells)
#1 COL1A1 / DCN  (mesenchymal cells) PDGFRA/PDGFRB
#2 SOX2 / NES / GFAP / S100B (glial?)
#3 ASCL1 / DLL1 / DLL3 / SOX2 / HOPX / MKI67(early progenitor)
#4 AQP4 / SLC1A2 / SOX2 / NES (astrocytes)
#5 MKI67 / TOP2A / ASCL1 / NES / SOX2 (proliferating NEPs)
#6 low quality
#7 COL1A1 / COL11A1 / MKI67 / TOP2A (proliferating mesenchymal cells)
#8 low quality

#save RDS
#saveRDS(adata_hu, "scRNA-seq_2023/Christensen_et_al_2024/data/rds/all_primary_human_NEPs_velocyte.rds")

```



_integrating our progenitor clusters from primary NEP cultures with the Sepp et al (PMID: 38029793) atlas_ 
The data has already been filtered, normalised, and 3x clustered (all features, gene features only, and TE only) as according to the README.
```{r}
#read in the data
sepp <- readRDS("../../../../shared/sharing_dir/snRNAseq_newborn_17_20_wpc_PMID_38029793/snRNAseq_newborn_17_20_wpc_PMID_38029793.Rds")

#filter out newborn data. 
Idents(sepp) <- sepp@meta.data$orig.ident
sepp <- subset(sepp, idents = c("17_wpc", "20_wpc"))

#assay back to RNA
sepp@active.assay <- "RNA"

#remove TE features
#extract the count matrix
counts <- sepp[["RNA"]]$counts
#identify the ones that are TEs. Select the ones that are not.
non_TE_index <- grep("^SoloTE", rownames(counts), invert = T)
#subset the count matrix
counts <- counts[non_TE_index,]
#make a subset of only the normal feature genes
sepp <- subset(sepp, features = rownames(counts))

#normalise data
sepp <- NormalizeData(sepp)

#perform CC scoring
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
sepp <- JoinLayers(sepp, overwrite = TRUE) #https://github.com/satijalab/seurat/issues/8304
sepp <- CellCycleScoring(sepp, s.features = s.genes, g2m.features = g2m.genes)

#add a column the difference between the S.score and G2M.Score to regress out cell cycle noise while still maintaining the difference between proliferative and non-proliferative cells at later stages
sepp@meta.data$CC.Difference <- sepp@meta.data$S.Score - sepp@meta.data$G2M.Score

#perform normalisation
sepp <- SCTransform(sepp, vars.to.regress = c("percent.mt", "CC.Difference", "nFeature_RNA", "nCount_RNA"))

#perform dimension reduction
sepp <- RunPCA(sepp)
sepp <- RunUMAP(sepp, dims = 1:30, reduction = "pca", return.model = TRUE, seed.use = 1234)

#look at QC values
VlnPlot(sepp, c("nCount_RNA", "nFeature_RNA", "percent.mt"), group.by = "orig.ident", pt.size = 1, alpha = 0.05)
FeaturePlot(sepp, c("nCount_RNA", "nFeature_RNA", "percent.mt")) & NoAxes()
DimPlot(sepp, group.by = c("orig.ident", "Phase")) & NoAxes()

# Filter based on nFeature_RNA and nCount_RNA 
sepp <- subset(sepp, subset = nFeature_RNA > 500 & nCount_RNA > 500)
VlnPlot(sepp, c("percent.mt", "CC.Difference", "nFeature_RNA", "nCount_RNA"))

#perform normalisation and dimension reduction again 
sepp <- SCTransform(sepp, vars.to.regress = c("percent.mt", "CC.Difference", "nFeature_RNA", "nCount_RNA"), conserve.memory = T)
sepp <- RunPCA(sepp)
sepp <- RunUMAP(sepp, dims = 1:30, reduction = "pca", return.model = TRUE, seed.use = 1234)

#look at QC values again
VlnPlot(sepp, c("nCount_RNA", "nFeature_RNA", "percent.mt"), group.by = "orig.ident", pt.size = 1, alpha = 0.05)
FeaturePlot(sepp, c("nCount_RNA", "nFeature_RNA", "percent.mt")) & NoAxes()
DimPlot(sepp, group.by = c("orig.ident", "Phase")) & NoAxes()

#perform clustering
sepp <- clusteringKit(sepp, assay="SCT", dim=20, res.range=seq(0.1, 2, 0.1), reduction="pca")

#plot to find the highest silhouette score w. low negative silhouette proportion
plot(sepp$sls_median~seq(0.1,2,0.1), type = "l", ylab = "Silhouette score") & plot(sepp$sls_neg_prop~seq(0.1,2,0.1), type = "l", ylab = "Negative silhouette proportion score")
#resolution of 0.3 is chosen based on biological meaningfulness

#rename the Seurat objects
sepp <- sepp$name

#Plot with the new resolutions
DimPlot(sepp, label = T, label.box = T, repel = T, group.by = "SCT_snn_res.0.3") + ggtitle("Sepp et al., pcw 17+20") & NoLegend() & NoAxes()
DimPlot(sepp, label = T, label.box = T, repel = T, group.by = "orig.ident") + ggtitle("Sepp et al., pcw 17+20") & NoLegend() & NoAxes()

#perform cluster analysis to identify marker genes in each cluster
DefaultAssay(sepp) <- 'SCT'
Idents(sepp) <- "SCT_snn_res.0.3"
sepp <- PrepSCTFindMarkers(sepp, assay = "SCT", verbose = TRUE)
sepp_markers <- FindAllMarkers(sepp, only.pos = TRUE, min.pct = 0.10, logfc.threshold = 0.25, assay = "SCT")

sepp_markers %>% dplyr::filter(cluster == 6) %>% transform(ratio = pct.1/pct.2) %>% arrange(desc(ratio))
sepp_markers %>% filter(gene == "SLC32A1")
FeaturePlot(sepp, features = c("GDF10", "GFAP", "NES"))


#0 VZ-derived progenitors (SOX2 / NES)
#1 GCPs (BARHL1 / RBFOX3 / PAX6)
#2 VZ-derived progenitors (SOX2 / NES)
#3 Early INs (ASCL1 / PAX2 / LHX5 / PTF1A)
#4 UBCs (EOMES / LMX1A)
#5 INs (GAD1 / GAD2)
#6 GCPs (BARHL1 / RBFOX3 / PAX6)
#7 PCs (SKOR2 / CALB1)
#8 Oligodendrocyte / OPCs (OLIG1 / OLIG2 / PDGFRA)
#9 Erythrocytes (HBB)
#10 microglia (SPP1 / CSF3R)


DotPlot(sepp, c("SOX2", "NES", "BARHL1", "RBFOX3", "PAX6", "ASCL1", "PAX2", "LHX5", "EOMES", "LMX1A", "GAD1", "GAD2", "SKOR2", "CALB1", "OLIG1", "OLIG2", "PDGFRA", "HBB", "CSF3R"), group.by = "SCT_snn_res.0.3")
```

_integration_
```{r}
#set the indent to the clustering with optimal resolution
Idents(adata_hu) <- "SCT_snn_res.0.3"

#the progenitor clusters of our in vitro cultures are subsetted for integration
integrator_rds_Hs <- subset(adata_hu, idents = c(2, 3, 4, 5))

#calculate top 200 variable features across that data set
marker_genes_sepp <- VariableFeatures(sepp, nfeatures = 200)
                  
#Subset the Sepp data to contain the top 200 variable features
sepp_subset <- sepp
DefaultAssay(sepp_subset) <- "SCT"
sepp_subset[["SCT"]] <- subset(sepp_subset[["SCT"]], features = marker_genes_sepp)


#identify transferanchors as described in https://satijalab.org/seurat/articles/integration_mapping
integrator_rds_Hs.anchors <- FindTransferAnchors(reference = sepp_subset, query = integrator_rds_Hs, dims = 1:40,
    reference.reduction = "pca", normalization.method = "SCT")


#use MapQuery to perform the label transfer as according to https://satijalab.org/seurat/articles/integration_mapping
integrator_rds_Hs <- MapQuery(anchorset = integrator_rds_Hs.anchors, reference = sepp_subset, query = integrator_rds_Hs,
    refdata = list(celltype = "SCT_snn_res.0.3"), reference.reduction = "pca", reduction.model = "umap", query.dims = 1000)


#alter the output column with predicted cell types for any label transfer with a prediction score less than 0.6
integrator_rds_Hs$celltype <- ifelse(integrator_rds_Hs$predicted.celltype.score > 0.6, integrator_rds_Hs$predicted.celltype, "Unassigned")

#plot the UMAP of the in vivo data and the in vitro data projected onto the reference UMAP
p1 <- DimPlot(sepp, reduction = "umap", group.by = "SCT_snn_res.0.3", label = F, label.size = 3,
    repel = TRUE) + ggtitle("Reference annotations") & NoAxes()
p2 <- DimPlot(integrator_rds_Hs, reduction = "ref.umap", group.by = "celltype", label = F,
    label.size = 3, repel = TRUE) + ggtitle("Query transferred labels")  & NoAxes() #, cols = my_color_palette
p1 + p2

#write a table with how big a percentage of the in vitro data is mapped to different clusters of the reference
table(integrator_rds_Hs@meta.data$celltype)/length(integrator_rds_Hs@meta.data$celltype)*100

#make a data from of the seurat object meta data to plot a sankey plot
df <- integrator_rds_Hs@meta.data %>% 
  ggsankey::make_long(sample, celltype)

df$node <- factor(df$node,levels = c("BRC_2584 (9w)","BRC_2583 (9w_4d)","BRC_2591 (10w_5d)", "BRC_2593 (12w)", "HDBR_18265 (12w)", "HDBR_18294 (12w)", "HDBR_18267 (13w)", "HDBR_18285 (13w)", "HDBR_18290 (15w)", "HDBR_18275 (17w)",
                                     "0", "1", "2", "3", "5", "6", "9", "10", "Unassigned"))

#plot a sankey plot - colours were changed in illustrator to match clustering
sankey_bs <- ggplot(df, aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               fill = node,
               label = node)) +
  geom_sankey(flow.alpha = 0.5, node.color = 1) +
  geom_sankey_label(size=2, fill = "white") +
  theme_sankey(base_size = 16) + NoLegend() +
  scale_x_discrete(labels=c(
    'sample' = 'In vitro cells',
    'celltype' = 'In vivo cells'
    ))  + labs(x = "")

#make a data from of the seurat object meta data to plot a sankey plot
df_by_cluster <- integrator_rds_Hs@meta.data %>% 
  ggsankey::make_long(SCT_snn_res.0.3, celltype)


#plot a sankey plot - colours were changed in illustrator to match clustering
sankey_bc <- ggplot(df_by_cluster, aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               fill = node,
               label = node)) +
  geom_sankey(flow.alpha = 0.5, node.color = 1) +
  geom_sankey_label(size=2, fill = "white") +
  theme_sankey(base_size = 16) + NoLegend() +
  scale_x_discrete(labels=c(
    'SCT_snn_res.0.3' = 'In vitro cells',
    'celltype' = 'In vivo cells'
    ))  + labs(x = "")

#visualise
sankey_bs + sankey_bc

# Extract UMAP embeddings
in_vivo_umap <- Embeddings(sepp, reduction = "umap") %>% as.data.frame()
in_vivo_umap$source <- "in_vivo"
in_vivo_umap$group <- sepp@meta.data$SCT_snn_res.0.4

in_vitro_umap <- Embeddings(integrator_rds_Hs, reduction = "ref.umap") %>% as.data.frame()
in_vitro_umap$source <- "in_vitro"
in_vitro_umap$group <- integrator_rds_Hs$celltype  # <- Replace with actual column

#provide the matching colour palette
library(scales)
my_color_palette <- hue_pal()(length(levels(as.factor(sepp@meta.data$SCT_snn_res.0.4))))
my_color_palette <- c(my_color_palette[1], my_color_palette[2],my_color_palette[3], my_color_palette[4], my_color_palette[6], my_color_palette[7], my_color_palette[10], my_color_palette[11], my_color_palette[13])


# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2), color = "grey80", size = 0.5) +
  geom_point(data = in_vitro_umap, aes(x = refUMAP_1, y = refUMAP_2, color = group), size = 0.7) +
  scale_color_manual(values = my_color_palette) +
  theme_void() +
  theme(legend.position = "right") & NoLegend()

# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2, color = group), size = 0.5) +
  scale_color_manual(values = hue_pal()(length(levels(as.factor(sepp@meta.data$SCT_snn_res.0.4))))) +
  theme_void() +
  theme(legend.position = "right")
```
