---
title: "In_vitro_integration"
output: html_document
date: '2025-04-26'
---

This is a script written by Jens Bager Christensen (jbc53@cam.ac.uk) to validate the cell identity of our primary NEP cultures established from neonatal mice cerebella.

Load packages and define functions
```{r}
library(Seurat)
library(tidyverse)
set.seed(1234)

#Defining parameters to test different resolutions for clustering
res.range <- c(seq(0.1,2,0.05))
assay <- "SCT"
dims <- 30
reduction <- "pca"

#Defining a function, which clusters a RDS in a range of resolutions, and calculates both the silhouette score and the negative silhouette proportion to determine which resolution is optimal
clusteringKit <- function(name, assay, dim, res.range, reduction){
  DefaultAssay(name) <- assay
  #using FindNeighbors (Seurat), a KNN based clustering algorithm, to compute a SNN graph, which then can be used to cluster the cells w. their neighbors
  name <- FindNeighbors(name, dims=1:dim, reduction=reduction)
  for (i in res.range){
    #FindClusters is used to cluster the aforementioned SNN graph by optimising the modularity function. This is carried out for all resolutions in the range of interest.
    name <- FindClusters(name, resolution=i)
  }
DefaultAssay(name) <- assay
  dist.matrix <- dist(x = Embeddings(object = name[[reduction]])[, 1:dim])
 clusters <- paste0(assay,"_snn_res.", res.range)
  getSil <- function(clr) {
    clr <- name@meta.data[[clr]]
    sil <- cluster::silhouette(x = as.numeric(x = as.factor(x = clr)), dist = dist.matrix)
    sil_value <- sil
    return(sil_value)
  }
  
  sls <- lapply(clusters, getSil) #mc.cores = 1)
  sls_median <- sapply(sls, function(x) median(x[,3])) %>% setNames(., res.range)
  sls_neg_prop <- sapply(sls, function(x) sum(x[,3]<0)/length(x[,3])) %>% setNames(., res.range)
  p_list <- lapply(res.range, function(res){
    #Idents(name) = name@meta.data[paste0(assay, "_snn_res.",res)]
    DimPlot(name, reduction = "umap", label = TRUE, group.by = paste0(assay, "_snn_res.",res), pt.size =0.3, raster=F)
  })
  return(list(name=name, sls_median=sls_median, sls_neg_prop=sls_neg_prop, dimplots=p_list))
}
```


_perform integration of primary mouse neurosphere cultures_

Read in the output of CellRanger aligned to GRCm39 and create a RDS object out of this. 
Classify cells based on the multiplex assay, which contains expression of the 12 used CMOs.
Thereafter, remove all _but_ CMO311 and CMO312 because these samples were mouse NEPs.
Following this vignette : https://www.10xgenomics.com/analysis-guides/tag-assignment-of-10x-Genomics-cellplex-data-using-seurats-htodemux-function
```{r}
#read in the output from 'cellranger multi' performed using CellRanger 8.0.1. Aligning only to GRCm39
data <- Read10X("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/primary_human_NEPs_GRCm39/outs/multi/count/raw_feature_bc_matrix/")

#Add the CellPlex as a CMO assay
adata = CreateSeuratObject(counts = data$`Gene Expression`)
adata[['CMO']] = CreateAssayObject(counts = data$`Multiplexing Capture`)

#identify cells with confident sample called based on cellranger
library(data.table)
cells <- fread("/mnt/beegfs6/home3/bayin/jbc53/CellRanger_CB/primary_human_NEPs_GRCm39/outs/multi/multiplexing_analysis/assignment_confidence_table.csv")

#remove cells assigned either Multiplet or Unassigned
cells <- cells %>% dplyr::filter(Assignment != "Multiplet" & Assignment != "Unassigned")

#extract only barcodes called with confident sample call
adata <- subset(adata, cells = cells$Barcode)

#remove cell without CMO counts
DefaultAssay(adata)<-"CMO"
adata <-  subset(x = adata, subset = nCount_CMO > 0)

#Normalize the CMO data and run the HTODemux function. Read more https://satijalab.org/seurat/archive/v3.1/hashing_vignette.html
adata <- NormalizeData(adata, assay = "CMO", normalization.method = "CLR")
adata <- HTODemux(adata, assay = "CMO", positive.quantile = 0.99)

#print out the amount of negative, doublets and singlets
print(table(adata$CMO_classification.global))

#Check the expression levels of CMO ID
#The true positive cells should be well distanced from the other CMO IDs 
RidgePlot(adata, assay = "CMO", features = rownames(adata[["CMO"]])[1:12], ncol = 6)

#Create individual Seurat objects of CMO311 and CMO3012, which are established from SW and Nes-CFP mice, respectively.
Idents(adata) <- "hash.ID"
adata_SW <- subset(adata, idents = c("CMO311"))
adata_NesCFP <- subset(adata, idents = c("CMO312"))
```


_The initial analysis is performed on the SW data starting with QC_

```{r}
#change assay to RNA
adata_SW@active.assay <- "RNA"

# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
adata_SW[["percent.mt"]] <- PercentageFeatureSet(adata_SW, pattern = "^mt-")

# Visualize QC metrics as a violin plot
VlnPlot(adata_SW, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, alpha = 0.05)

#remove outliers beased on amount of features detected and too high mitochondrial reads
adata_SW <- subset(adata_SW, subset = nFeature_RNA > 200 & nFeature_RNA < 7500 & percent.mt < 7.5)

#run sctransform, normalising features, counts and percent mitochondrial reads
adata_SW <- SCTransform(adata_SW, vars.to.regress = c("nFeature_RNA", "nCount_RNA", "percent.mt"), verbose = FALSE)

#perform dimension reduction using PCA and UMAP to visualise the data 
adata_SW <- RunPCA(adata_SW, verbose = FALSE)
adata_SW <- RunUMAP(adata_SW, dims = 1:30, verbose = FALSE)

#perform cell cycle scoring and regress out the difference between S and G2M
m.cc.genes <- lapply(cc.genes, str_to_title)
adata_SW <- CellCycleScoring(adata_SW, s.features = m.cc.genes$s.genes, g2m.features = m.cc.genes$g2m.genes)

#visualise the phase on a DimPlot
DimPlot(adata_SW, group.by = "Phase")

#regress out the difference between G2M and S using sctransform
adata_SW$CC.Difference <- adata_SW$S.Score - adata_SW$G2M.Score
adata_SW <- SCTransform(adata_SW, vars.to.regress = c("CC.Difference"))

#redo the dimension reduction
adata_SW <- RunPCA(adata_SW, verbose = FALSE)
adata_SW <- RunUMAP(adata_SW, dims = 1:30, verbose = FALSE)

#visualise the Dimplot with metadata variables of interest
DimPlot(adata_SW, group.by = "hash.ID") & NoAxes()
DimPlot(adata_SW, group.by = "Phase") & NoAxes()

#featureplot markers of interest
FeaturePlot(adata_SW, features = c("Nes", "Sox2", "Ascl1", "Hopx", "Slc6a11", "Foxo1")) & NoAxes()
```


_Perform clustering and cluster analysis to annotate the clusters_
```{r}
#try to cluster using a range of resolutions to identifying a meaningful resolution parameter
adata_SW <- clusteringKit(adata_SW, assay="SCT", dim=30, res.range=seq(0.1, 3, 0.1), reduction="pca")

#plot the Silhouette score and negative Silhouette score to get an impression of the different clustering resolutions
plot(adata_SW$sls_median~seq(0.1,3,0.1), type = "l", ylab = "Silhouette score") & plot(adata_SW$sls_neg_prop~seq(0.1,3,0.1), type = "l", ylab = "Negative silhouette proportion score")
#resolution of 0.2 seems most appropriate 

#rename the object
adata_SW <- adata_SW$name

#Plot with the identified resolution and compare to the previous annotations
DimPlot(adata_SW, label = T, label.box = T, repel = T, group.by = "SCT_snn_res.0.2") + ggtitle("Mouse cells clustered with a resolution of 0.2") & NoLegend() & NoAxes()

#calculate marker genes for the clusters by setting the assay to RNA and the idents to the clustering computed with the optimal resolution.
DefaultAssay(adata_SW) <- 'SCT'
Idents(adata_SW) <- "SCT_snn_res.0.2"
adata_SW <- PrepSCTFindMarkers(adata_SW, assay = "SCT", verbose = TRUE)
adata_SW_markers <- FindAllMarkers(adata_SW, only.pos = TRUE, min.pct = 0.10, logfc.threshold = 0.25, assay = "SCT")

#save the markers
#write_csv(adata_SW_markers, "Ms_in_vitro/SW_cluster_markers.csv")

#Cluster annotations
#0 Hopx / Nes / Sox2 (Hopx-NEPs)
#1 Mki67 / Top2a / Nes / Hopx / Ascl1 (proliferating NEPs)
#2 Ascl1 / Nes (Ascl1-NEPs)

#featureplot markers of interest
VlnPlot(adata_SW, features = c("Ascl1", "Nes", "Hopx", "Sox2", "Crlf1", "Slc6a11"), group.by = "SCT_snn_res.0.2")
DotPlot(adata_SW, features = c("Hopx", "Nes", "Sox2", "Mki67", "Top2a", "Ascl1"), group.by = "SCT_snn_res.0.2")

FeaturePlot(adata_SW, features = c("Hopx", "Nes", "Sox2", "Mki67", "Top2a", "Ascl1"))

#save the suerat object
#saveRDS(adata_SW, "../data/rds/primary_human_NEPs_SW.rds")
```


_Integrating with in vivo data_
Starting off by integrating the in vitro data with our full in vivo data set.
```{r}
#require("AnnotationDbi")
#require("org.Mm.eg.db")
#require("TimeSeriesExperiment")
#require(ggsankey)
#require(org.Mm.eg.db)

#back up the seurat rds in case the process does not work out
integrator_rds <- adata_SW

#Read in the full (nonIR) data set from the in vivo experiment
in_vivo <- readRDS(file = "../data/rds/nonIR_all_NEPs.rds")

#calculate top 200 variable features across that data set
marker_genes_in_vivo <- VariableFeatures(in_vivo, nfeatures = 200)


#because the in vitro data has gene symbols instead of ENSEMBL IDs we transform the RDS to a single cell experiment object
#to rename the genes to be compatible with the in vivo genes
integrator_rds <- as.SingleCellExperiment(integrator_rds, assay = "SCT")

#use mapIds and the org.Mm.eg.db to make a dictionary with gene symbols and ENSEMBL IDs
gene_names <- mapIds(org.Mm.eg.db, keys=rownames(integrator_rds), keytype="SYMBOL", columns="ENSEMBL",column="ENSEMBL")

#add the ENSEMBL IDs of the gene names as rowData of the sce object
rowData(integrator_rds)$ENSEMBL <- gene_names
#add the gene symbols of the gene names as rowData of the sce object
rowData(integrator_rds)$SYMBOL <- rownames(integrator_rds)

#remove the ones where no ENSEMBL ID could be found (e.g. NAs)
integrator_rds <- integrator_rds[! is.na(rowData(integrator_rds)$ENSEMBL),]

#make the gene list unique
integrator_rds <- integrator_rds[isUnique(rowData(integrator_rds)$ENSEMBL),]

#subset only the genes which are part of the computed top 200 variable features from the in vivo data
#this is done to ensure meaningful genes are used for the label transfer and it is not driven by e.g. lowly expressed genes
integrator_rds <- integrator_rds[rowData(integrator_rds)$ENSEMBL %in% marker_genes_in_vivo,]

#rename the genes of the in vitro data to the ENSEMBL counterparts
row.names(integrator_rds) <- rowData(integrator_rds)$ENSEMBL

#convert the sce object back into a seurat object
integrator_rds <- as.Seurat(integrator_rds, counts = "counts", data = "logcounts")

#identify transferanchors as described in https://satijalab.org/seurat/articles/integration_mapping
integrator_rds.anchors <- FindTransferAnchors(reference = in_vivo, query = integrator_rds, dims = 1:40,
    reference.reduction = "pca", normalization.method = "SCT")

#set seed for consistency
set.seed(1234)
#redo the UMAP reductino of the in vivo data, because return.model needs to be TRUE for the MapQuery
in_vivo <- RunUMAP(in_vivo, dims = 1:30, n.neighbors = 30, return.model = T)


#use MapQuery to perform the label transfer as according to https://satijalab.org/seurat/articles/integration_mapping
integrator_rds <- MapQuery(anchorset = integrator_rds.anchors, reference = in_vivo, query = integrator_rds,
    refdata = list(celltype = "SCT_snn_res.0.4"), reference.reduction = "pca", reduction.model = "umap", query.dims = 1000)


#alter the output column with predicted cell types for any label transfer with a prediction score less than 0.6
integrator_rds$celltype <- ifelse(integrator_rds$predicted.celltype.score > 0.6, integrator_rds$predicted.celltype, "Unassigned")

#provide the matching colour palette
library(scales)
my_color_palette <- hue_pal()(length(levels(as.factor(in_vivo@meta.data$SCT_snn_res.0.4))))
my_color_palette <- c(my_color_palette[1], my_color_palette[3],my_color_palette[5], my_color_palette[7],my_color_palette[12])

#plot the UMAP of the in vivo data and the in vitro data projected onto the reference UMAP
p1 <- DimPlot(in_vivo, reduction = "umap", group.by = "SCT_snn_res.0.4", label = F, label.size = 3,
    repel = TRUE) + ggtitle("Reference annotations") + xlim(-20,20) + ylim(-20,20) & NoAxes()
p2 <- DimPlot(integrator_rds, reduction = "ref.umap", group.by = "celltype", label = F,
    label.size = 3, repel = TRUE) + ggtitle("Query transferred labels")  + xlim(-20,20) + ylim(-20,20) & NoAxes() #, cols = my_color_palette
p1 + p2

#write a table with how big a percentage of the in vitro data is mapped to different clusters of the reference
table(integrator_rds@meta.data$celltype)/length(integrator_rds@meta.data$celltype)*100

#make a data from of the seurat object meta data to plot a sankey plot
df <- integrator_rds@meta.data %>% 
  ggsankey::make_long(CMO_maxID, celltype)

#plot a sankey plot - colours were changed in illustrator to match clustering
ggplot(df, aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               fill = node,
               label = node)) +
  geom_sankey(flow.alpha = 0.5, node.color = 1) +
  geom_sankey_label(size=2, fill = "white") +
  theme_sankey(base_size = 16) + NoLegend() +
  scale_x_discrete(labels=c(
    'CMO_maxID' = 'In vitro cells',
    'celltype' = 'In vivo cells'
    ))  + labs(x = "")


# Extract UMAP embedding
in_vivo_umap <- Embeddings(in_vivo, reduction = "umap") %>% as.data.frame()
in_vivo_umap$source <- "in_vivo"
in_vivo_umap$group <- in_vivo@meta.data$SCT_snn_res.0.4

in_vitro_umap <- Embeddings(integrator_rds, reduction = "ref.umap") %>% as.data.frame()
in_vitro_umap$source <- "in_vitro"
in_vitro_umap$group <- integrator_rds$celltype  # <- Replace with actual column

# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2), color = "grey80", size = 0.5) +
  geom_point(data = in_vitro_umap, aes(x = refUMAP_1, y = refUMAP_2, color = group), size = 0.7) +
  scale_color_manual(values = my_color_palette) +
  theme_void() +
  theme(legend.position = "right") & NoLegend()

# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2, color = group), size = 0.5) +
  scale_color_manual(values = hue_pal()(length(levels(as.factor(in_vivo@meta.data$SCT_snn_res.0.4))))) +
  theme_void() +
  theme(legend.position = "right")
```

Secondly, we try to integrate it with our NEPs-only subset
```{r}
#back up the seurat rds in case the process does not work out
integrator_rds <- adata_SW

#Read in the full (nonIR) data set from the in vivo experiment
in_vivo_NEPs <- readRDS(file = "../data/rds/nonIR_VZ_NEPs.rds")

#calculate top 200 variable features across that data set
marker_genes_in_vivo_NEPs <- VariableFeatures(in_vivo_NEPs, nfeatures = 200)


#because the in vitro data has gene symbols instead of ENSEMBL IDs we transform the RDS to a single cell experiment object
#to rename the genes to be compatible with the in vivo genes
integrator_rds <- as.SingleCellExperiment(integrator_rds, assay = "SCT")

#use mapIds and the org.Mm.eg.db to make a dictionary with gene symbols and ENSEMBL IDs
gene_names <- mapIds(org.Mm.eg.db, keys=rownames(integrator_rds), keytype="SYMBOL", columns="ENSEMBL",column="ENSEMBL")

#add the ENSEMBL IDs of the gene names as rowData of the sce object
rowData(integrator_rds)$ENSEMBL <- gene_names
#add the gene symbols of the gene names as rowData of the sce object
rowData(integrator_rds)$SYMBOL <- rownames(integrator_rds)

#remove the ones where no ENSEMBL ID could be found (e.g. NAs)
integrator_rds <- integrator_rds[! is.na(rowData(integrator_rds)$ENSEMBL),]

#make the gene list unique
integrator_rds <- integrator_rds[isUnique(rowData(integrator_rds)$ENSEMBL),]

#subset only the genes which are part of the computed top 200 variable features from the in vivo data
#this is done to ensure meaningful genes are used for the label transfer and it is not driven by e.g. lowly expressed genes
integrator_rds <- integrator_rds[rowData(integrator_rds)$ENSEMBL %in% marker_genes_in_vivo_NEPs,]

#rename the genes of the in vitro data to the ENSEMBL counterparts
row.names(integrator_rds) <- rowData(integrator_rds)$ENSEMBL

#convert the sce object back into a seurat object
integrator_rds <- as.Seurat(integrator_rds, counts = "counts", data = "logcounts")

#identify transferanchors as described in https://satijalab.org/seurat/articles/integration_mapping
integrator_rds.anchors <- FindTransferAnchors(reference = in_vivo_NEPs, query = integrator_rds, dims = 1:40,
    reference.reduction = "pca", normalization.method = "SCT")

#set seed for consistency
set.seed(1234)
#redo the UMAP reductino of the in vivo data, because return.model needs to be TRUE for the MapQuery
in_vivo_NEPs <- RunUMAP(in_vivo_NEPs, dims = 1:30, n.neighbors = 30, return.model = T)


#use MapQuery to perform the label transfer as according to https://satijalab.org/seurat/articles/integration_mapping
integrator_rds <- MapQuery(anchorset = integrator_rds.anchors, reference = in_vivo_NEPs, query = integrator_rds,
    refdata = list(celltype = "SCT_snn_res.0.9"), reference.reduction = "pca", reduction.model = "umap", query.dims = 1000)


#alter the output column with predicted cell types for any label transfer with a prediction score less than 0.6
integrator_rds$celltype <- ifelse(integrator_rds$predicted.celltype.score > 0.5, integrator_rds$predicted.celltype, "Unassigned")

#plot the UMAP of the in vivo data and the in vitro data projected onto the reference UMAP
p1 <- DimPlot(in_vivo_NEPs, reduction = "umap", group.by = "SCT_snn_res.0.9", label = F, label.size = 3,
    repel = TRUE) + ggtitle("Reference annotations") + xlim(-20,20) + ylim(-20,20) & NoAxes()
p2 <- DimPlot(integrator_rds, reduction = "ref.umap", group.by = "celltype", label = F,
    label.size = 3, repel = TRUE) + ggtitle("Query transferred labels")  + xlim(-20,20) + ylim(-20,20) & NoAxes() #, cols = my_color_palette
p1 + p2

#write a table with how big a percentage of the in vitro data is mapped to different clusters of the reference
table(integrator_rds@meta.data$celltype)/length(integrator_rds@meta.data$celltype)*100

#make a data from of the seurat object meta data to plot a sankey plot
df <- integrator_rds@meta.data %>% 
  ggsankey::make_long(CMO_maxID, celltype)

#plot a sankey plot - colours were changed in illustrator to match clustering
ggplot(df, aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               fill = node,
               label = node)) +
  geom_sankey(flow.alpha = 0.5, node.color = 1) +
  geom_sankey_label(size=2, fill = "white") +
  theme_sankey(base_size = 16) + NoLegend() +
  scale_x_discrete(labels=c(
    'CMO_maxID' = 'In vitro cells',
    'celltype' = 'In vivo cells'
    ))  + labs(x = "")


# Extract UMAP embedding
in_vivo_umap <- Embeddings(in_vivo_NEPs, reduction = "umap") %>% as.data.frame()
in_vivo_umap$source <- "in_vivo"
in_vivo_umap$group <- in_vivo_NEPs@meta.data$SCT_snn_res.0.9

in_vitro_umap <- Embeddings(integrator_rds, reduction = "ref.umap") %>% as.data.frame()
in_vitro_umap$source <- "in_vitro"
in_vitro_umap$group <- integrator_rds$celltype

# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2), color = "grey80", size = 0.5) +
  geom_point(data = in_vitro_umap, aes(x = refUMAP_1, y = refUMAP_2, color = group), size = 0.7) +
  #scale_color_manual(values = my_color_palette) +
  theme_void() +
  theme(legend.position = "right") & NoLegend()

# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2, color = group), size = 0.5) +
  scale_color_manual(values = hue_pal()(length(levels(as.factor(in_vivo_NEPs@meta.data$SCT_snn_res.0.9))))) +
  theme_void() +
  theme(legend.position = "right")
```




_similarly the same analysis is perform on the cell derived from the NES-CFP animal_

```{r}
#change assay to RNA
adata_NesCFP@active.assay <- "RNA"

# The [[ operator can add columns to object metadata. This is a great place to stash QC stats
adata_NesCFP[["percent.mt"]] <- PercentageFeatureSet(adata_NesCFP, pattern = "^mt-")

# Visualize QC metrics as a violin plot
VlnPlot(adata_NesCFP, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, alpha = 0.05)

#remove outliers beased on amount of features detected and too high mitochondrial reads
adata_NesCFP <- subset(adata_NesCFP, subset = nFeature_RNA > 200 & nFeature_RNA < 7500 & percent.mt < 7.5)

#run sctransform, normalising features, counts and percent mitochondrial reads
adata_NesCFP <- SCTransform(adata_NesCFP, vars.to.regress = c("nFeature_RNA", "nCount_RNA", "percent.mt"), verbose = FALSE)

#perform dimension reduction using PCA and UMAP to visualise the data 
adata_NesCFP <- RunPCA(adata_NesCFP, verbose = FALSE)
adata_NesCFP <- RunUMAP(adata_NesCFP, dims = 1:30, verbose = FALSE)

#perform cell cycle scoring and regress out the difference between S and G2M
adata_NesCFP <- CellCycleScoring(adata_NesCFP, s.features = m.cc.genes$s.genes, g2m.features = m.cc.genes$g2m.genes)

#visualise the phase on a DimPlot
DimPlot(adata_NesCFP, group.by = "Phase")

#regress out the difference between G2M and S using sctransform
adata_NesCFP$CC.Difference <- adata_NesCFP$S.Score - adata_NesCFP$G2M.Score
adata_NesCFP <- SCTransform(adata_NesCFP, vars.to.regress = c("CC.Difference"))

#redo the dimension reduction
adata_NesCFP <- RunPCA(adata_NesCFP, verbose = FALSE)
adata_NesCFP <- RunUMAP(adata_NesCFP, dims = 1:30, verbose = FALSE)

#visualise the Dimplot with metadata variables of interest
DimPlot(adata_NesCFP, group.by = "hash.ID") & NoAxes()
DimPlot(adata_NesCFP, group.by = "Phase") & NoAxes()

#featureplot markers of interest
FeaturePlot(adata_NesCFP, features = c("Nes", "Sox2", "Ascl1", "Hopx", "Slc6a11", "Foxo1")) & NoAxes()
```

_Perform clustering and cluster analysis to annotate the clusters_
```{r}
#try to cluster using a range of resolutions to identifying a meaningful resolution parameter
adata_NesCFP <- clusteringKit(adata_NesCFP, assay="SCT", dim=30, res.range=seq(0.1, 3, 0.1), reduction="pca")

#plot the Silhouette score and negative Silhouette score to get an impression of the different clustering resolutions
plot(adata_NesCFP$sls_median~seq(0.1,3,0.1), type = "l", ylab = "Silhouette score") & plot(adata_NesCFP$sls_neg_prop~seq(0.1,3,0.1), type = "l", ylab = "Negative silhouette proportion score")
#resolution of 0.3 seems most appropriate 

#rename the object
adata_NesCFP <- adata_NesCFP$name

#Plot with the identified resolution and compare to the previous annotations
DimPlot(adata_NesCFP, label = T, label.box = T, repel = T, group.by = "SCT_snn_res.0.3") + ggtitle("Mouse cells clustered with a resolution of 0.3") & NoLegend() & NoAxes()

#calculate marker genes for the clusters by setting the assay to RNA and the idents to the clustering computed with the optimal resolution.
DefaultAssay(adata_NesCFP) <- 'SCT'
Idents(adata_NesCFP) <- "SCT_snn_res.0.3"
adata_NesCFP <- PrepSCTFindMarkers(adata_NesCFP, assay = "SCT", verbose = TRUE)
adata_NesCFP_markers <- FindAllMarkers(adata_NesCFP, only.pos = TRUE, min.pct = 0.10, logfc.threshold = 0.25, assay = "SCT")

#save the markers
write_csv(adata_NesCFP_markers, "Ms_in_vitro/NesCFP_cluster_markers.csv")

#Cluster annotations
#0 Ascl1 / Nes
#1 Hopx / Sox2 / Nes
#2 Mki67 / Top2a

#featureplot markers of interest
VlnPlot(adata_NesCFP, features = c("Ascl1", "Nes", "Hopx", "Sox2", "Crlf1", "Slc6a11"), group.by = "SCT_snn_res.0.2")
DotPlot(adata_NesCFP, features = c("Hopx", "Nes", "Sox2", "Mki67", "Top2a", "Ascl1"), group.by = "SCT_snn_res.0.2")

FeaturePlot(adata_NesCFP, features = c("Hopx", "Nes", "Sox2", "Mki67", "Top2a", "Ascl1"))

#save sueart object
#saveRDS(adata_NesCFP, "../data/rds/primary_human_NEPs_NesCFP.rds")
```




_Integrating with in vivo data_
Like for the primary SW cultures, we start by integrating it with our full in vivo data set
```{r}
#require("AnnotationDbi")
#require("org.Mm.eg.db")
#require("TimeSeriesExperiment")
#require(ggsankey)
#require(org.Mm.eg.db)

#back up the seurat rds in case the process does not work out
integrator_rds <- adata_NesCFP

#Read in the full (nonIR) data set from the in vivo experiment
in_vivo <- readRDS(file = "../data/rds/nonIR_all_NEPs.rds")

#calculate top 200 variable features across that data set
marker_genes_in_vivo <- VariableFeatures(in_vivo, nfeatures = 200)


#because the in vitro data has gene symbols instead of ENSEMBL IDs we transform the RDS to a single cell experiment object
#to rename the genes to be compatible with the in vivo genes
integrator_rds <- as.SingleCellExperiment(integrator_rds, assay = "SCT")

#use mapIds and the org.Mm.eg.db to make a dictionary with gene symbols and ENSEMBL IDs
gene_names <- mapIds(org.Mm.eg.db, keys=rownames(integrator_rds), keytype="SYMBOL", columns="ENSEMBL",column="ENSEMBL")

#add the ENSEMBL IDs of the gene names as rowData of the sce object
rowData(integrator_rds)$ENSEMBL <- gene_names
#add the gene symbols of the gene names as rowData of the sce object
rowData(integrator_rds)$SYMBOL <- rownames(integrator_rds)

#remove the ones where no ENSEMBL ID could be found (e.g. NAs)
integrator_rds <- integrator_rds[! is.na(rowData(integrator_rds)$ENSEMBL),]

#make the gene list unique
integrator_rds <- integrator_rds[isUnique(rowData(integrator_rds)$ENSEMBL),]

#subset only the genes which are part of the computed top 200 variable features from the in vivo data
#this is done to ensure meaningful genes are used for the label transfer and it is not driven by e.g. lowly expressed genes
integrator_rds <- integrator_rds[rowData(integrator_rds)$ENSEMBL %in% marker_genes_in_vivo,]

#rename the genes of the in vitro data to the ENSEMBL counterparts
row.names(integrator_rds) <- rowData(integrator_rds)$ENSEMBL

#convert the sce object back into a seurat object
integrator_rds <- as.Seurat(integrator_rds, counts = "counts", data = "logcounts")

#identify transferanchors as described in https://satijalab.org/seurat/articles/integration_mapping
integrator_rds.anchors <- FindTransferAnchors(reference = in_vivo, query = integrator_rds, dims = 1:40,
    reference.reduction = "pca", normalization.method = "SCT")

#set seed for consistency
set.seed(1234)
#redo the UMAP reductino of the in vivo data, because return.model needs to be TRUE for the MapQuery
in_vivo <- RunUMAP(in_vivo, dims = 1:30, n.neighbors = 30, return.model = T)


#use MapQuery to perform the label transfer as according to https://satijalab.org/seurat/articles/integration_mapping
integrator_rds <- MapQuery(anchorset = integrator_rds.anchors, reference = in_vivo, query = integrator_rds,
    refdata = list(celltype = "SCT_snn_res.0.4"), reference.reduction = "pca", reduction.model = "umap", query.dims = 1000)


#alter the output column with predicted cell types for any label transfer with a prediction score less than 0.6
integrator_rds$celltype <- ifelse(integrator_rds$predicted.celltype.score > 0.6, integrator_rds$predicted.celltype, "Unassigned")

#provide the matching colour palette
library(scales)
my_color_palette <- hue_pal()(length(levels(as.factor(in_vivo@meta.data$SCT_snn_res.0.4))))
my_color_palette <- c(my_color_palette[1], my_color_palette[5], my_color_palette[7],my_color_palette[12])

#plot the UMAP of the in vivo data and the in vitro data projected onto the reference UMAP
p1 <- DimPlot(in_vivo, reduction = "umap", group.by = "SCT_snn_res.0.4", label = F, label.size = 3,
    repel = TRUE) + ggtitle("Reference annotations") + xlim(-20,20) + ylim(-20,20) & NoAxes()
p2 <- DimPlot(integrator_rds, reduction = "ref.umap", group.by = "celltype", label = F,
    label.size = 3, repel = TRUE) + ggtitle("Query transferred labels")  + xlim(-20,20) + ylim(-20,20) & NoAxes() #, cols = my_color_palette
p1 + p2

#write a table with how big a percentage of the in vitro data is mapped to different clusters of the reference
table(integrator_rds@meta.data$celltype)/length(integrator_rds@meta.data$celltype)*100

#make a data from of the seurat object meta data to plot a sankey plot
df <- integrator_rds@meta.data %>% 
  ggsankey::make_long(CMO_maxID, celltype)

#plot a sankey plot - colours were changed in illustrator to match clustering
ggplot(df, aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               fill = node,
               label = node)) +
  geom_sankey(flow.alpha = 0.5, node.color = 1) +
  geom_sankey_label(size=2, fill = "white") +
  theme_sankey(base_size = 16) + NoLegend() +
  scale_x_discrete(labels=c(
    'CMO_maxID' = 'In vitro cells',
    'celltype' = 'In vivo cells'
    ))  + labs(x = "")


# Extract UMAP embedding
in_vivo_umap <- Embeddings(in_vivo, reduction = "umap") %>% as.data.frame()
in_vivo_umap$source <- "in_vivo"
in_vivo_umap$group <- in_vivo@meta.data$SCT_snn_res.0.4

in_vitro_umap <- Embeddings(integrator_rds, reduction = "ref.umap") %>% as.data.frame()
in_vitro_umap$source <- "in_vitro"
in_vitro_umap$group <- integrator_rds$celltype  # <- Replace with actual column

# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2), color = "grey80", size = 0.5) +
  geom_point(data = in_vitro_umap, aes(x = refUMAP_1, y = refUMAP_2, color = group), size = 0.7) +
  scale_color_manual(values = my_color_palette) +
  theme_void() +
  theme(legend.position = "right") & NoLegend()

# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2, color = group), size = 0.5) +
  scale_color_manual(values = hue_pal()(length(levels(as.factor(in_vivo@meta.data$SCT_snn_res.0.4))))) +
  theme_void() +
  theme(legend.position = "right")
```


Secondly, we try to integrate it with our NEPs-only subset
```{r}
#back up the seurat rds in case the process does not work out
integrator_rds <- adata_NesCFP

#Read in the full (nonIR) data set from the in vivo experiment
in_vivo_NEPs <- readRDS(file = "../data/rds/nonIR_VZ_NEPs.rds")

#calculate top 200 variable features across that data set
marker_genes_in_vivo_NEPs <- VariableFeatures(in_vivo_NEPs, nfeatures = 200)


#because the in vitro data has gene symbols instead of ENSEMBL IDs we transform the RDS to a single cell experiment object
#to rename the genes to be compatible with the in vivo genes
integrator_rds <- as.SingleCellExperiment(integrator_rds, assay = "SCT")

#use mapIds and the org.Mm.eg.db to make a dictionary with gene symbols and ENSEMBL IDs
gene_names <- mapIds(org.Mm.eg.db, keys=rownames(integrator_rds), keytype="SYMBOL", columns="ENSEMBL",column="ENSEMBL")

#add the ENSEMBL IDs of the gene names as rowData of the sce object
rowData(integrator_rds)$ENSEMBL <- gene_names
#add the gene symbols of the gene names as rowData of the sce object
rowData(integrator_rds)$SYMBOL <- rownames(integrator_rds)

#remove the ones where no ENSEMBL ID could be found (e.g. NAs)
integrator_rds <- integrator_rds[! is.na(rowData(integrator_rds)$ENSEMBL),]

#make the gene list unique
integrator_rds <- integrator_rds[isUnique(rowData(integrator_rds)$ENSEMBL),]

#subset only the genes which are part of the computed top 200 variable features from the in vivo data
#this is done to ensure meaningful genes are used for the label transfer and it is not driven by e.g. lowly expressed genes
integrator_rds <- integrator_rds[rowData(integrator_rds)$ENSEMBL %in% marker_genes_in_vivo_NEPs,]

#rename the genes of the in vitro data to the ENSEMBL counterparts
row.names(integrator_rds) <- rowData(integrator_rds)$ENSEMBL

#convert the sce object back into a seurat object
integrator_rds <- as.Seurat(integrator_rds, counts = "counts", data = "logcounts")

#identify transferanchors as described in https://satijalab.org/seurat/articles/integration_mapping
integrator_rds.anchors <- FindTransferAnchors(reference = in_vivo_NEPs, query = integrator_rds, dims = 1:40,
    reference.reduction = "pca", normalization.method = "SCT")

#set seed for consistency
set.seed(1234)
#redo the UMAP reductino of the in vivo data, because return.model needs to be TRUE for the MapQuery
in_vivo_NEPs <- RunUMAP(in_vivo_NEPs, dims = 1:30, n.neighbors = 30, return.model = T)


#use MapQuery to perform the label transfer as according to https://satijalab.org/seurat/articles/integration_mapping
integrator_rds <- MapQuery(anchorset = integrator_rds.anchors, reference = in_vivo_NEPs, query = integrator_rds,
    refdata = list(celltype = "SCT_snn_res.0.9"), reference.reduction = "pca", reduction.model = "umap", query.dims = 1000)


#alter the output column with predicted cell types for any label transfer with a prediction score less than 0.6
integrator_rds$celltype <- ifelse(integrator_rds$predicted.celltype.score > 0.5, integrator_rds$predicted.celltype, "Unassigned")

#plot the UMAP of the in vivo data and the in vitro data projected onto the reference UMAP
p1 <- DimPlot(in_vivo_NEPs, reduction = "umap", group.by = "SCT_snn_res.0.9", label = F, label.size = 3,
    repel = TRUE) + ggtitle("Reference annotations") + xlim(-20,20) + ylim(-20,20) & NoAxes()
p2 <- DimPlot(integrator_rds, reduction = "ref.umap", group.by = "celltype", label = F,
    label.size = 3, repel = TRUE) + ggtitle("Query transferred labels")  + xlim(-20,20) + ylim(-20,20) & NoAxes() #, cols = my_color_palette
p1 + p2

#write a table with how big a percentage of the in vitro data is mapped to different clusters of the reference
table(integrator_rds@meta.data$celltype)/length(integrator_rds@meta.data$celltype)*100

#make a data from of the seurat object meta data to plot a sankey plot
df <- integrator_rds@meta.data %>% 
  ggsankey::make_long(CMO_maxID, celltype)

#plot a sankey plot - colours were changed in illustrator to match clustering
ggplot(df, aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               fill = node,
               label = node)) +
  geom_sankey(flow.alpha = 0.5, node.color = 1) +
  geom_sankey_label(size=2, fill = "white") +
  theme_sankey(base_size = 16) + NoLegend() +
  scale_x_discrete(labels=c(
    'CMO_maxID' = 'In vitro cells',
    'celltype' = 'In vivo cells'
    ))  + labs(x = "")


# Extract UMAP embedding
in_vivo_umap <- Embeddings(in_vivo_NEPs, reduction = "umap") %>% as.data.frame()
in_vivo_umap$source <- "in_vivo"
in_vivo_umap$group <- in_vivo_NEPs@meta.data$SCT_snn_res.0.9

in_vitro_umap <- Embeddings(integrator_rds, reduction = "ref.umap") %>% as.data.frame()
in_vitro_umap$source <- "in_vitro"
in_vitro_umap$group <- integrator_rds$celltype

# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2), color = "grey80", size = 0.5) +
  geom_point(data = in_vitro_umap, aes(x = refUMAP_1, y = refUMAP_2, color = group), size = 0.7) +
  #scale_color_manual(values = my_color_palette) +
  theme_void() +
  theme(legend.position = "right") & NoLegend()

# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2, color = group), size = 0.5) +
  scale_color_manual(values = hue_pal()(length(levels(as.factor(in_vivo_NEPs@meta.data$SCT_snn_res.0.9))))) +
  theme_void() +
  theme(legend.position = "right")
```
































_sex regression work in progress_
```{r}
#see that sex seemingly is a big driver in the differences in the two samples based on Xist expression
DimPlot(adata_ms, group.by = "hash.ID")
FeaturePlot(adata_ms, features = "Xist")

#Add a module score for each cell based on their Xist expression. This is necessary to regress out the sex difference.
adata_ms <- AddModuleScore(object = adata_ms, features = "Xist", name = "Xist", ctrl = 5)


#include the new Xist module score in SCTransform to regress out the effect
adata_ms <- SCTransform(adata_ms, vars.to.regress = c("Xist1","CC.Difference"))

#redo the dimension reduction
adata_ms <- RunPCA(adata_ms, verbose = FALSE)
adata_ms <- RunUMAP(adata_ms, dims = 1:30, verbose = FALSE)

#visualise
DimPlot(adata_ms, group.by = "hash.ID") & NoAxes()
FeaturePlot(adata_ms, c("Xist", "Ddx3y", "Eif2s3y", "Uty", "Kdm5d"))
```





_Integrating with in vivo data_
```{r}
#require("AnnotationDbi")
#require("org.Mm.eg.db")
#require("TimeSeriesExperiment")
#require(ggsankey)

#back up the seurat rds in case the process does not work out
integrator_rds <- adata_ms

#Read in the full (nonIR) data set from the in vivo experiment
in_vivo <- readRDS(file = "../data/rds/nonIR_all_NEPs.rds")

#calculate top 200 variable features across that data set
marker_genes_in_vivo <- VariableFeatures(in_vivo, nfeatures = 200)


#because the in vitro data has gene symbols instead of ENSEMBL IDs we transform the RDS to a single cell experiment object
#to rename the genes to be compatible with the in vivo genes
integrator_rds <- as.SingleCellExperiment(integrator_rds, assay = "SCT")

#use mapIds and the org.Mm.eg.db to make a dictionary with gene symbols and ENSEMBL IDs
gene_names <- mapIds(org.Mm.eg.db, keys=rownames(integrator_rds), keytype="SYMBOL", columns="ENSEMBL",column="ENSEMBL")

#add the ENSEMBL IDs of the gene names as rowData of the sce object
rowData(integrator_rds)$ENSEMBL <- gene_names
#add the gene symbols of the gene names as rowData of the sce object
rowData(integrator_rds)$SYMBOL <- rownames(integrator_rds)

#remove the ones where no ENSEMBL ID could be found (e.g. NAs)
integrator_rds <- integrator_rds[! is.na(rowData(integrator_rds)$ENSEMBL),]

#make the gene list unique
integrator_rds <- integrator_rds[isUnique(rowData(integrator_rds)$ENSEMBL),]

#subset only the genes which are part of the computed top 200 variable features from the in vivo data
#this is done to ensure meaningful genes are used for the label transfer and it is not driven by e.g. lowly expressed genes
integrator_rds <- integrator_rds[rowData(integrator_rds)$ENSEMBL %in% marker_genes_in_vivo,]

#rename the genes of the in vitro data to the ENSEMBL counterparts
row.names(integrator_rds) <- rowData(integrator_rds)$ENSEMBL

#convert the sce object back into a seurat object
integrator_rds <- as.Seurat(integrator_rds, counts = "counts", data = "logcounts")

#identify transferanchors as described in https://satijalab.org/seurat/articles/integration_mapping
integrator_rds.anchors <- FindTransferAnchors(reference = in_vivo, query = integrator_rds, dims = 1:40,
    reference.reduction = "pca", normalization.method = "SCT")

set.seed(1234)
#redo the UMAP reductino of the in vivo data, because return.model needs to be TRUE for the MapQuery
in_vivo <- RunUMAP(in_vivo, dims = 1:30, n.neighbors = 30, return.model = T)


#use MapQuery to perform the label transfer as according to https://satijalab.org/seurat/articles/integration_mapping
integrator_rds <- MapQuery(anchorset = integrator_rds.anchors, reference = in_vivo, query = integrator_rds,
    refdata = list(celltype = "SCT_snn_res.0.4"), reference.reduction = "pca", reduction.model = "umap", query.dims = 1000)


#alter the output column with predicted cell types for any label transfer with a prediction score less than 0.6
integrator_rds$celltype <- ifelse(integrator_rds$predicted.celltype.score > 0.6, integrator_rds$predicted.celltype, "Unassigned")

#provide the matching colour palette
library(scales)
my_color_palette <- hue_pal()(length(levels(as.factor(in_vivo@meta.data$SCT_snn_res.0.4))))
my_color_palette <- c(my_color_palette[1], my_color_palette[5],my_color_palette[7], my_color_palette[12])

#plot the UMAP of the in vivo data and the in vitro data projected onto the reference UMAP
p1 <- DimPlot(in_vivo, reduction = "umap", group.by = "SCT_snn_res.0.4", label = F, label.size = 3,
    repel = TRUE) + ggtitle("Reference annotations") + xlim(-20,20) + ylim(-20,20) & NoAxes()
p2 <- DimPlot(integrator_rds, reduction = "ref.umap", group.by = "celltype", label = F,
    label.size = 3, repel = TRUE) + ggtitle("Query transferred labels")  + xlim(-20,20) + ylim(-20,20) & NoAxes() #, cols = my_color_palette
p1 + p2

#write a table with how big a percentage of the in vitro data is mapped to different clusters of the reference
table(integrator_rds@meta.data$celltype)/length(integrator_rds@meta.data$celltype)*100

#make a data from of the seurat object meta data to plot a sankey plot
df <- integrator_rds@meta.data %>% 
  ggsankey::make_long(CMO_maxID, celltype)

#plot a sankey plot - colours were changed in illustrator to match clustering
ggplot(df, aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               fill = node,
               label = node)) +
  geom_sankey(flow.alpha = 0.5, node.color = 1) +
  geom_sankey_label(size=2, fill = "white") +
  theme_sankey(base_size = 16) + NoLegend() +
  scale_x_discrete(labels=c(
    'CMO_maxID' = 'In vitro cells',
    'celltype' = 'In vivo cells'
    ))  + labs(x = "")


# Extract UMAP embeddings
in_vivo_umap <- Embeddings(in_vivo, reduction = "umap") %>% as.data.frame()
in_vivo_umap$source <- "in_vivo"
in_vivo_umap$group <- in_vivo@meta.data$SCT_snn_res.0.4

in_vitro_umap <- Embeddings(integrator_rds, reduction = "ref.umap") %>% as.data.frame()
in_vitro_umap$source <- "in_vitro"
in_vitro_umap$group <- integrator_rds$celltype  # <- Replace with actual column

# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2), color = "grey80", size = 0.5) +
  geom_point(data = in_vitro_umap, aes(x = refUMAP_1, y = refUMAP_2, color = group), size = 0.7) +
  scale_color_manual(values = my_color_palette) +
  theme_void() +
  theme(legend.position = "right") & NoLegend()

# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2, color = group), size = 0.5) +
  scale_color_manual(values = hue_pal()(length(levels(as.factor(in_vivo@meta.data$SCT_snn_res.0.4))))) +
  theme_void() +
  theme(legend.position = "right")
```





















































_integrating in vitro data with in vivo data_
```{r}
require("AnnotationDbi")
require("org.Mm.eg.db")
require("TimeSeriesExperiment")
require(ggsankey)

#back up the seurat rds in case the process does not work out
integrator_rds <- adata_ms

#Read in the full (nonIR) data set from the in vivo experiment
in_vivo <- readRDS(file = "scRNA-seq_2023/Christensen_et_al_2024/data/rds/nonIR_all_NEPs.rds")

#calculate top 200 variable features across that data set
#marker_genes_in_vivo <- VariableFeatures(in_vivo, nfeatures = 200)
#alternatively the genes used to annotate can be used
marker_genes_in_vivo <- c("Hopx", "Gdf10", "Ascl1", "Ptf1a", "Barhl1", "Atoh1", "Rbfox3", "Pax2", "Tubb3", "Slc6a11", "Barhl1", "Atoh1", "Foxj1", "Otx2", "Vtn", "Dcn", "Olig2", "Gpr17", "Pdgfra", "Ly68", "Cx3cr1")

#because the in vitro data has gene symbols instead of ENSEMBL IDs we transform the RDS to a single cell experiment object
#to rename the genes to be compatible with the in vivo genes
integrator_rds <- as.SingleCellExperiment(integrator_rds, assay = "SCT")

#use mapIds and the org.Mm.eg.db to make a dictionary with gene symbols and ENSEMBL IDs
gene_names <- mapIds(org.Mm.eg.db, keys=rownames(integrator_rds), keytype="SYMBOL", columns="ENSEMBL",column="ENSEMBL")

#add the ENSEMBL IDs of the gene names as rowData of the sce object
rowData(integrator_rds)$ENSEMBL <- gene_names
#add the gene symbols of the gene names as rowData of the sce object
rowData(integrator_rds)$SYMBOL <- rownames(integrator_rds)

#remove the ones where no ENSEMBL ID could be found (e.g. NAs)
integrator_rds <- integrator_rds[! is.na(rowData(integrator_rds)$ENSEMBL),]

#make the gene list unique
integrator_rds <- integrator_rds[isUnique(rowData(integrator_rds)$ENSEMBL),]

#subset only the genes which are part of the computed top 200 variable features from the in vivo data
#this is done to ensure meaningful genes are used for the label transfer and it is not driven by e.g. lowly expressed genes
#integrator_rds <- integrator_rds[rowData(integrator_rds)$ENSEMBL %in% marker_genes_in_vivo,]
integrator_rds <- integrator_rds[rowData(integrator_rds)$SYMBOL %in% marker_genes_in_vivo,]

#rename the genes of the in vitro data to the ENSEMBL counterparts
row.names(integrator_rds) <- rowData(integrator_rds)$ENSEMBL

#convert the sce object back into a seurat object
integrator_rds <- as.Seurat(integrator_rds, counts = "counts", data = "logcounts")

#identify transferanchors as described in https://satijalab.org/seurat/articles/integration_mapping
integrator_rds.anchors <- FindTransferAnchors(reference = in_vivo, query = integrator_rds, dims = 1:40,
    reference.reduction = "pca", normalization.method = "SCT")

set.seed(1234)
#redo the UMAP reductino of the in vivo data, because return.model needs to be TRUE for the MapQuery
in_vivo <- RunUMAP(in_vivo, dims = 1:30, n.neighbors = 30, return.model = T)


#use MapQuery to perform the label transfer as according to https://satijalab.org/seurat/articles/integration_mapping
#integrator_rds <- MapQuery(anchorset = integrator_rds.anchors, reference = in_vivo, query = integrator_rds,
#    refdata = list(celltype = "nonIR_annotations"), reference.reduction = "pca", reduction.model = "umap", query.dims = 1000)
integrator_rds <- MapQuery(anchorset = integrator_rds.anchors, reference = in_vivo, query = integrator_rds,
    refdata = list(celltype = "SCT_snn_res.0.4"), reference.reduction = "pca", reduction.model = "umap", query.dims = 1000)


#alter the output column with predicted cell types for any label transfer with a prediction score less than 0.6
integrator_rds$celltype <- ifelse(integrator_rds$predicted.celltype.score > 0.6, integrator_rds$predicted.celltype, "Unassigned")

#provide the matching colour palette
library(scales)
my_color_palette <- hue_pal()(length(levels(as.factor(in_vivo@meta.data$SCT_snn_res.0.4))))
my_color_palette <- c(my_color_palette[1:2], my_color_palette[5:6], my_color_palette[12])

#plot the UMAP of the in vivo data and the in vitro data projected onto the reference UMAP
p1 <- DimPlot(in_vivo, reduction = "umap", group.by = "SCT_snn_res.0.4", label = F, label.size = 3,
    repel = TRUE) + ggtitle("Reference annotations") + xlim(-20,20) + ylim(-20,20) & NoAxes()
p2 <- DimPlot(integrator_rds, reduction = "ref.umap", group.by = "celltype", label = F,
    label.size = 3, repel = TRUE, cols = my_color_palette) + ggtitle("Query transferred labels")  + xlim(-20,20) + ylim(-20,20) & NoAxes()
p1 + p2

#write a table with how big a percentage of the in vitro data is mapped to different clusters of the reference
table(integrator_rds@meta.data$celltype)/length(integrator_rds@meta.data$celltype)*100

#make a data from of the seurat object meta data to plot a sankey plot
df <- integrator_rds@meta.data %>% 
  ggsankey::make_long(CMO_maxID, celltype)

#plot a sankey plot - colours were changed in illustrator to match clustering
ggplot(df, aes(x = x, 
               next_x = next_x, 
               node = node, 
               next_node = next_node,
               fill = node,
               label = node)) +
  geom_sankey(flow.alpha = 0.5, node.color = 1) +
  geom_sankey_label(size=2, fill = "white") +
  theme_sankey(base_size = 16) + NoLegend() +
  scale_x_discrete(labels=c(
    'CMO_maxID' = 'In vitro cells',
    'celltype' = 'In vivo cells'
    ))  + labs(x = "")


# Extract UMAP embeddings
in_vivo_umap <- Embeddings(in_vivo, reduction = "umap") %>% as.data.frame()
in_vivo_umap$source <- "in_vivo"
in_vivo_umap$group <- in_vivo@meta.data$SCT_snn_res.0.4

in_vitro_umap <- Embeddings(integrator_rds, reduction = "ref.umap") %>% as.data.frame()
in_vitro_umap$source <- "in_vitro"
in_vitro_umap$group <- integrator_rds$celltype  # <- Replace with actual column

# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2), color = "grey80", size = 0.5) +
  geom_point(data = in_vitro_umap, aes(x = refUMAP_1, y = refUMAP_2, color = group), size = 0.7) +
  scale_color_manual(values = my_color_palette) +
  theme_void() +
  theme(legend.position = "right")

# Plot
ggplot() +
  geom_point(data = in_vivo_umap, aes(x = umap_1, y = umap_2, color = group), size = 0.5) +
  scale_color_manual(values = hue_pal()(length(levels(as.factor(in_vivo@meta.data$SCT_snn_res.0.4))))) +
  theme_void() +
  theme(legend.position = "right")
```
