---
title: "R Notebook"
output: html_notebook
---

This is a script written by Jens Bager Christensen (jbc53@cam.ac.uk) to gain insight into potential DNA binding proteins co-binding with ASCL1 and FOXO1 genes during NEP differentiation.
A list of genes containing both a significant ASCL1 and FOXO1 peak either ±1kb or ±2kb of its TSS are computed. This gene list is then used to filter down peaks from the Dam-alone condition. The analysis is performed both at day 2 and day 6.


Day 2
±1kb
```{r}
library(ChIPpeakAnno)
#load the significant peaks called by Alex
ASCL1_d2 <- toGRanges("/mnt/beegfs6/home3/bayin/jbc53/DamID/ASCL1-FOXO1_DamID/significant_peaks/repro_peaks/ASCL1_D2_5.reproPeak_nohead.bed.txt", format = "BED", header = T, )
FOXO1_d2 <- toGRanges("/mnt/beegfs6/home3/bayin/jbc53/DamID/ASCL1-FOXO1_DamID/significant_peaks/repro_peaks/FOXO_D2.5.reproPeak_nohead.bed.txt", format = "BED", header = T)

#load the UCSC mm10 genome (the same the sequencing was mapped to)
library(EnsDb.Mmusculus.v79)

EnsDb.Mmusculus.v79.genes <- genes(EnsDb.Mmusculus.v79)

#use biomaRt to add gene symbols instead of ensembl IDs
library(biomaRt)


mart <- useMart(biomart="ENSEMBL_MART_ENSEMBL", 
                dataset = "mmusculus_gene_ensembl", 
                host = "https://www.ensembl.org/", verbose = F)

#annotate
FOXO1_d2 <- annotatePeakInBatch(FOXO1_d2, 
                                         AnnotationData = EnsDb.Mmusculus.v79.genes)

ASCL1_d2 <- annotatePeakInBatch(ASCL1_d2, 
                                         AnnotationData = EnsDb.Mmusculus.v79.genes)

#remove peaks that are more than 1000 bases from a gene TSS
FOXO1_d2 <- FOXO1_d2[FOXO1_d2$distancetoFeature < 1000]
FOXO1_d2 <- FOXO1_d2[FOXO1_d2$distancetoFeature > -1000]
#remove peaks that are more than 1000 bases from a gene TSS
ASCL1_d2<- ASCL1_d2[ASCL1_d2$distancetoFeature < 1000]
ASCL1_d2 <- ASCL1_d2[ASCL1_d2$distancetoFeature > -1000]

#unique features
# Subset the GRanges object to include only the first occurrence of each unique feature
FOXO1_d2_uni_feat <- FOXO1_d2[!duplicated(mcols(FOXO1_d2)$feature)]
ASCL1_d2_uni_feat <- ASCL1_d2[!duplicated(mcols(ASCL1_d2)$feature)]

library(ggvenn)
#make a list of the genes to create a Venndiagram
venn_list <- list(ASCL1_d2 = ASCL1_d2_uni_feat$feature, 
     FOXO1_d2 = FOXO1_d2_uni_feat$feature)

#plot
ggvenn(venn_list, auto_scale = T, fill_alpha = 0.5)

#find genes having a significant peak ± 1 kb from the TSS from ASCL1 day 2 and FOXO1 day 2
shared_genes <- intersect(ASCL1_d2_uni_feat$feature, FOXO1_d2_uni_feat$feature)



#read in Dam-alone day 2
Dam_d2 <- toGRanges("/mnt/beegfs6/home3/bayin/jbc53/DamID/ASCL1-FOXO1_DamID/significant_peaks/repro_peaks/DAM_D2_5.reproPeak_no_header.bed", format = "BED", header = T)

#annotate 
Dam_d2 <- annotatePeakInBatch(Dam_d2, 
                                         AnnotationData = EnsDb.Mmusculus.v79.genes)

#subset only peak nearest genes who has both ASCL1 and FOXO1 peak ±1kb of their TSS at day2
Dam_d2_subset <- Dam_d2[Dam_d2$feature %in% shared_genes]

#reformat for homer
Dam_d2_subset_homer <- data.frame(chr=seqnames(Dam_d2_subset),
  start=start(Dam_d2_subset)-1,
  end=end(Dam_d2_subset),
  name=elementMetadata(Dam_d2_subset)$feature,
  strand=strand(Dam_d2_subset))

Dam_d2_subset_homer <- Dam_d2_subset_homer %>% transform(chr = paste0("chr", chr))

Dam_d2_subset_homer <- Dam_d2_subset_homer %>% transform(ID = 1:length(Dam_d2_subset_homer$start)) %>% dplyr::select(c("chr", "start", "end", "ID", "name", "strand"))

#write out the bed-file for HOMER analysis
write.table(Dam_d2_subset_homer, file="/mnt/home3/bayin/jbc53/DamID/In_vitro_NEP_DamID/homer/data/Dam-alone_d2_shared_genes_1kb.bed", quote=F, sep="\t", row.names=F, col.names=F)

#correct HOMER output for multiple testing by Bonferroni
d2_1kb <- read.csv("DamID/In_vitro_NEP_DamID/homer/day2_shared_1kb/knownResults.txt", sep = "\t")

d2_1kb <- d2_1kb %>% transform(p.adj = P.value * length(d2_1kb$Motif.Name))

d2_1kb <- d2_1kb %>% mutate(p.adj = pmin(p.adj, 1))

#write out the correct output as an Excel
write.xlsx(d2_1kb, "DamID/In_vitro_NEP_DamID/homer/day2_shared_1kb/KnownResults_corrected.xlsx")
```


Day 2
±2kb
```{r}
library(ChIPpeakAnno)
#load the significant peaks called by Alex
ASCL1_d2 <- toGRanges("/mnt/beegfs6/home3/bayin/jbc53/DamID/ASCL1-FOXO1_DamID/significant_peaks/repro_peaks/ASCL1_D2_5.reproPeak_nohead.bed.txt", format = "BED", header = T, )
FOXO1_d2 <- toGRanges("/mnt/beegfs6/home3/bayin/jbc53/DamID/ASCL1-FOXO1_DamID/significant_peaks/repro_peaks/FOXO_D2.5.reproPeak_nohead.bed.txt", format = "BED", header = T)

#load the UCSC mm10 genome (the same the sequencing was mapped to)
library(EnsDb.Mmusculus.v79)

EnsDb.Mmusculus.v79.genes <- genes(EnsDb.Mmusculus.v79)

#use biomaRt to add gene symbols instead of ensembl IDs
library(biomaRt)


mart <- useMart(biomart="ENSEMBL_MART_ENSEMBL", 
                dataset = "mmusculus_gene_ensembl", 
                host = "https://www.ensembl.org/", verbose = F)

#annotate
FOXO1_d2 <- annotatePeakInBatch(FOXO1_d2, 
                                         AnnotationData = EnsDb.Mmusculus.v79.genes)

ASCL1_d2 <- annotatePeakInBatch(ASCL1_d2, 
                                         AnnotationData = EnsDb.Mmusculus.v79.genes)

#remove peaks that are more than 1000 bases from a gene TSS
FOXO1_d2 <- FOXO1_d2[FOXO1_d2$distancetoFeature < 2000]
FOXO1_d2 <- FOXO1_d2[FOXO1_d2$distancetoFeature > -2000]
#remove peaks that are more than 1000 bases from a gene TSS
ASCL1_d2<- ASCL1_d2[ASCL1_d2$distancetoFeature < 2000]
ASCL1_d2 <- ASCL1_d2[ASCL1_d2$distancetoFeature > -2000]

#unique features
# Subset the GRanges object to include only the first occurrence of each unique feature
FOXO1_d2_uni_feat <- FOXO1_d2[!duplicated(mcols(FOXO1_d2)$feature)]
ASCL1_d2_uni_feat <- ASCL1_d2[!duplicated(mcols(ASCL1_d2)$feature)]

library(ggvenn)
#make a list of the genes to create a Venndiagram
venn_list <- list(ASCL1_d2 = ASCL1_d2_uni_feat$feature, 
     FOXO1_d2 = FOXO1_d2_uni_feat$feature)

#plot
ggvenn(venn_list, auto_scale = T, fill_alpha = 0.5)

#find genes having a significant peak ± 2 kb from the TSS from ASCL1 day 2 and FOXO1 day 2
shared_genes <- intersect(ASCL1_d2_uni_feat$feature, FOXO1_d2_uni_feat$feature)



#read in Dam-alone day 2
Dam_d2 <- toGRanges("/mnt/beegfs6/home3/bayin/jbc53/DamID/ASCL1-FOXO1_DamID/significant_peaks/repro_peaks/DAM_D2_5.reproPeak_no_header.bed", format = "BED", header = T)

#annotate 
Dam_d2 <- annotatePeakInBatch(Dam_d2, 
                                         AnnotationData = EnsDb.Mmusculus.v79.genes)

#subset only peak nearest genes who has both ASCL1 and FOXO1 peak ±2kb of their TSS at day2
Dam_d2_subset <- Dam_d2[Dam_d2$feature %in% shared_genes]

#reformat for HOMER analysis
Dam_d2_subset_homer <- data.frame(chr=seqnames(Dam_d2_subset),
  start=start(Dam_d2_subset)-1,
  end=end(Dam_d2_subset),
  name=elementMetadata(Dam_d2_subset)$feature,
  strand=strand(Dam_d2_subset))

Dam_d2_subset_homer <- Dam_d2_subset_homer %>% transform(chr = paste0("chr", chr))

Dam_d2_subset_homer <- Dam_d2_subset_homer %>% transform(ID = 1:length(Dam_d2_subset_homer$start)) %>% dplyr::select(c("chr", "start", "end", "ID", "name", "strand"))

#write out the bed-file
write.table(Dam_d2_subset_homer, file="/mnt/home3/bayin/jbc53/DamID/In_vitro_NEP_DamID/homer/data/Dam-alone_d2_shared_genes_2kb.bed", quote=F, sep="\t", row.names=F, col.names=F)

#correct HOMER output for multiple testing by Bonferroni
d2_2kb <- read.csv("DamID/In_vitro_NEP_DamID/homer/day2_shared_2kb/knownResults.txt", sep = "\t")

d2_2kb <- d2_2kb %>% transform(p.adj = P.value * length(d2_2kb$Motif.Name))

d2_2kb <- d2_2kb %>% mutate(p.adj = pmin(p.adj, 1))

#write out the output as an Excel
write.xlsx(d2_2kb, "DamID/In_vitro_NEP_DamID/homer/day2_shared_2kb/KnownResults_corrected.xlsx")
```



Day 6
±1kb
```{r}
library(ChIPpeakAnno)
#load the significant peaks called by Alex
ASCL1_d6 <- toGRanges("/mnt/beegfs6/home3/bayin/jbc53/DamID/ASCL1-FOXO1_DamID/significant_peaks/repro_peaks/ASCL1_D6_5.reproPeak_nohead.bed.txt", format = "BED", header = T, )
FOXO1_d6 <- toGRanges("/mnt/beegfs6/home3/bayin/jbc53/DamID/ASCL1-FOXO1_DamID/significant_peaks/repro_peaks/FOXO1_D6_5.reproPeak_nohead.bed.txt", format = "BED", header = T)

#load the UCSC mm10 genome (the same the sequencing was mapped to)
library(EnsDb.Mmusculus.v79)

EnsDb.Mmusculus.v79.genes <- genes(EnsDb.Mmusculus.v79)

#use biomaRt to add gene symbols instead of ensembl IDs
library(biomaRt)


mart <- useMart(biomart="ENSEMBL_MART_ENSEMBL", 
                dataset = "mmusculus_gene_ensembl", 
                host = "https://www.ensembl.org/", verbose = F)

#annotate
FOXO1_d6 <- annotatePeakInBatch(FOXO1_d6, 
                                         AnnotationData = EnsDb.Mmusculus.v79.genes)

ASCL1_d6 <- annotatePeakInBatch(ASCL1_d6, 
                                         AnnotationData = EnsDb.Mmusculus.v79.genes)

#remove peaks that are more than 1000 bases from a gene TSS
FOXO1_d6 <- FOXO1_d6[FOXO1_d6$distancetoFeature < 1000]
FOXO1_d6 <- FOXO1_d6[FOXO1_d6$distancetoFeature > -1000]
#remove peaks that are more than 1000 bases from a gene TSS
ASCL1_d6<- ASCL1_d6[ASCL1_d6$distancetoFeature < 1000]
ASCL1_d6 <- ASCL1_d6[ASCL1_d6$distancetoFeature > -1000]

#unique features
# Subset the GRanges object to include only the first occurrence of each unique feature
FOXO1_d6_uni_feat <- FOXO1_d6[!duplicated(mcols(FOXO1_d6)$feature)]
ASCL1_d6_uni_feat <- ASCL1_d6[!duplicated(mcols(ASCL1_d6)$feature)]

library(ggvenn)
#make a list of the genes to create a Venndiagram
venn_list <- list(ASCL1_d6 = ASCL1_d6_uni_feat$feature, 
     FOXO1_d6 = FOXO1_d6_uni_feat$feature)

#plot
ggvenn(venn_list, auto_scale = T, fill_alpha = 0.5)

#find genes having a significant peak ± 1 kb from the TSS from ASCL1 day 2 and FOXO1 day 2
shared_genes <- intersect(ASCL1_d6_uni_feat$feature, FOXO1_d6_uni_feat$feature)



#read in Dam-alone day 6
Dam_d6 <- toGRanges("/mnt/beegfs6/home3/bayin/jbc53/DamID/ASCL1-FOXO1_DamID/significant_peaks/repro_peaks/DAM_D6_5.reproPeak_no_header.bed", format = "BED", header = T)

#annotate 
Dam_d6 <- annotatePeakInBatch(Dam_d6, 
                                         AnnotationData = EnsDb.Mmusculus.v79.genes)

#subset only peak nearest genes who has both ASCL1 and FOXO1 peak ±1kb of their TSS at day6
Dam_d6_subset <- Dam_d6[Dam_d6$feature %in% shared_genes]

#reformat for HOMER analysis
Dam_d6_subset_homer <- data.frame(chr=seqnames(Dam_d6_subset),
  start=start(Dam_d6_subset)-1,
  end=end(Dam_d6_subset),
  name=elementMetadata(Dam_d6_subset)$feature,
  strand=strand(Dam_d6_subset))

Dam_d6_subset_homer <- Dam_d6_subset_homer %>% transform(chr = paste0("chr", chr))

Dam_d6_subset_homer <- Dam_d6_subset_homer %>% transform(ID = 1:length(Dam_d6_subset_homer$start)) %>% dplyr::select(c("chr", "start", "end", "ID", "name", "strand"))

#write out the bed-file
write.table(Dam_d6_subset_homer, file="/mnt/home3/bayin/jbc53/DamID/In_vitro_NEP_DamID/homer/data/Dam-alone_d6_shared_genes_1kb.bed", quote=F, sep="\t", row.names=F, col.names=F)

#correct HOMER output for multiple testing by Bonferroni
d6_1kb <- read.csv("DamID/In_vitro_NEP_DamID/homer/day6_shared_1kb/knownResults.txt", sep = "\t")

d6_1kb <- d6_1kb %>% transform(p.adj = P.value * length(d6_1kb$Motif.Name))

d6_2kb <- d6_2kb %>% mutate(p.adj = pmin(p.adj, 1))

#write out the corrected output as an Excel
write.xlsx(d6_1kb, "DamID/In_vitro_NEP_DamID/homer/day6_shared_1kb/KnownResults_corrected.xlsx")
```


Day 6
±2kb
```{r}
library(ChIPpeakAnno)
#load the significant peaks called by Alex
ASCL1_d6 <- toGRanges("/mnt/beegfs6/home3/bayin/jbc53/DamID/ASCL1-FOXO1_DamID/significant_peaks/repro_peaks/ASCL1_D6_5.reproPeak_nohead.bed.txt", format = "BED", header = T, )
FOXO1_d6 <- toGRanges("/mnt/beegfs6/home3/bayin/jbc53/DamID/ASCL1-FOXO1_DamID/significant_peaks/repro_peaks/FOXO1_D6_5.reproPeak_nohead.bed.txt", format = "BED", header = T)

#load the UCSC mm10 genome (the same the sequencing was mapped to)
library(EnsDb.Mmusculus.v79)

EnsDb.Mmusculus.v79.genes <- genes(EnsDb.Mmusculus.v79)

#use biomaRt to add gene symbols instead of ensembl IDs
library(biomaRt)


mart <- useMart(biomart="ENSEMBL_MART_ENSEMBL", 
                dataset = "mmusculus_gene_ensembl", 
                host = "https://www.ensembl.org/", verbose = F)

#annotate
FOXO1_d6 <- annotatePeakInBatch(FOXO1_d6, 
                                         AnnotationData = EnsDb.Mmusculus.v79.genes)

ASCL1_d6 <- annotatePeakInBatch(ASCL1_d6, 
                                         AnnotationData = EnsDb.Mmusculus.v79.genes)

#remove peaks that are more than 2000 bases from a gene TSS
FOXO1_d6 <- FOXO1_d6[FOXO1_d6$distancetoFeature < 2000]
FOXO1_d6 <- FOXO1_d6[FOXO1_d6$distancetoFeature > -2000]
#remove peaks that are more than 2000 bases from a gene TSS
ASCL1_d6<- ASCL1_d6[ASCL1_d6$distancetoFeature < 2000]
ASCL1_d6 <- ASCL1_d6[ASCL1_d6$distancetoFeature > -2000]

#unique features
# Subset the GRanges object to include only the first occurrence of each unique feature
FOXO1_d6_uni_feat <- FOXO1_d6[!duplicated(mcols(FOXO1_d6)$feature)]
ASCL1_d6_uni_feat <- ASCL1_d6[!duplicated(mcols(ASCL1_d6)$feature)]

library(ggvenn)
#make a list of the genes to create a Venndiagram
venn_list <- list(ASCL1_d6 = ASCL1_d6_uni_feat$feature, 
     FOXO1_d6 = FOXO1_d6_uni_feat$feature)

#plot
ggvenn(venn_list, auto_scale = T, fill_alpha = 0.5)

#find genes having a significant peak 2 1 kb from the TSS from ASCL1 day 2 and FOXO1 day 6
shared_genes <- intersect(ASCL1_d6_uni_feat$feature, FOXO1_d6_uni_feat$feature)



#read in Dam-alone day 6
Dam_d6 <- toGRanges("/mnt/beegfs6/home3/bayin/jbc53/DamID/ASCL1-FOXO1_DamID/significant_peaks/repro_peaks/DAM_D6_5.reproPeak_no_header.bed", format = "BED", header = T)

#annotate 
Dam_d6 <- annotatePeakInBatch(Dam_d6, 
                                         AnnotationData = EnsDb.Mmusculus.v79.genes)

#subset only peak nearest genes who has both ASCL1 and FOXO1 peak ±2kb of their TSS at day6
Dam_d6_subset <- Dam_d6[Dam_d6$feature %in% shared_genes]

#reformat for HOMER analysis
Dam_d6_subset_homer <- data.frame(chr=seqnames(Dam_d6_subset),
  start=start(Dam_d6_subset)-1,
  end=end(Dam_d6_subset),
  name=elementMetadata(Dam_d6_subset)$feature,
  strand=strand(Dam_d6_subset))

Dam_d6_subset_homer <- Dam_d6_subset_homer %>% transform(chr = paste0("chr", chr))

Dam_d6_subset_homer <- Dam_d6_subset_homer %>% transform(ID = 1:length(Dam_d6_subset_homer$start)) %>% dplyr::select(c("chr", "start", "end", "ID", "name", "strand"))

#write out the BED-file
write.table(Dam_d6_subset_homer, file="/mnt/home3/bayin/jbc53/DamID/In_vitro_NEP_DamID/homer/data/Dam-alone_d6_shared_genes_2kb.bed", quote=F, sep="\t", row.names=F, col.names=F)

#correct HOMER output for multiple testing by Bonferroni
d6_2kb <- read.csv("DamID/In_vitro_NEP_DamID/homer/day6_shared_2kb/knownResults.txt", sep = "\t")

d6_2kb <- d6_2kb %>% transform(p.adj = P.value * length(d6_2kb$Motif.Name))

d6_2kb <- d6_2kb %>% mutate(p.adj = pmin(p.adj, 1))

#write out the corrected output as an Excel
write.xlsx(d6_2kb, "DamID/In_vitro_NEP_DamID/homer/day6_shared_2kb/KnownResults_corrected.xlsx")
```




